---
title: "Data Divas Midterm"
author: "Simintaj Salehpour, Madeline Bumpus, Nazish Atta, Divya Mistry"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(skimr)
library(naniar)
library(dplyr)
library(ezids)
```

### Examining the Relationship Between Demographics, Admission Types, and Clinical Outcomes in New York Inpatients (Niagara County)

## 1. Introduction

**Due:** October 22nd (Wednesday, End of Day)

This R Markdown performs an exploratory data analysis (EDA) for the dataset `Hospital_Inpatient_Discharges.csv`. It contains code, explanation, dataset summary, descriptive statistics, graphical displays, variance/SD measures, normality checks, and initial statistical tests. 

## 2. Load dataset
```{r Load Dataset, include=T, results='markup'}
# Load CSV file
data <- read.csv("Hospital_Inpatient_Discharges.csv")

#Clean column names to avoid backticks

names(data) <- make.names(names(data))

cat("Dataset dimensions: Rows:", nrow(data), "Columns:", ncol(data), "\n")
glimpse(data)

```

## 3. Handle missing value

Here I check if there is a missing value or not. For each column check the missing value.

```{r Missing Value, include=T, results = "markup"}

# --- 1. Summary of missing values ---
missing_summary <- data %>% summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "column(variable)", values_to = "n_missing") %>% arrange(desc(n_missing))

missing_summary

# --- 2. Visualize missingness per column ---
gg_miss_var(data) + labs(title = "Variables VS Number of Missing Values") +
  theme( plot.title = element_text(face = "bold", color = "darkblue"))

# --- 3. Check if there are any missing values ---
clean_data <- data
if(any(is.na(data))) {
  
  cat("Missing values detected!\n")
  total_missing <- sum(is.na(data))
  total_cells <- nrow(data) * ncol(data)
  
  # Remove rows if missing values are few (<5% of total data)
  if(total_missing / total_cells < 0.05) {
    clean_data <- na.omit(data)
    cat("Rows with missing values removed.\n")
  } 
  else {
    # Impute numeric columns with median
    for(col in names(clean_data)) {
      if(is.numeric(clean_data[[col]])) {          # Check if the column is numeric
          clean_data[[col]][is.na(clean_data[[col]])] <- median(clean_data[[col]], na.rm = TRUE)  
      }
    }
    cat("Missing values in numeric columns replaced with median.\n")
  }
  
} else {
  cat("No missing values detected. Dataset is clean.\n")
}


```

## 4. Remove Duplicates
Here I check to remove duplicates:
```{r Remove Duplicates, include=T, results = "markup"}
dup_count <- sum(duplicated(clean_data))
cat("Number of duplicate rows:", dup_count, "\n")

# Remove duplicates
clean_data <- clean_data %>% distinct()
```

## 5. Descriptive statistics

```{r descriptive-stats, include=T, results = "markup"}
# For numeric columns
cat("Numeric Columns summary:")
num_vars <- clean_data %>% select(where(is.numeric))
skim(num_vars)


# Categorical Columns Summary (top counts for first 5 columns)
cat_vars <- clean_data %>% select(where(~!is.numeric(.)))
cols_to_show <- names(cat_vars)[1:min(5, ncol(cat_vars))]

cat("Categorical Columns Summary (first", length(cols_to_show), "columns):\n\n")

for(col in cols_to_show) {
  cat("Column:", col, "\n")
  counts <- sort(table(cat_vars[[col]]), decreasing = TRUE)
  print(head(counts, 6))  # top 6 categories
  cat("\n")
}

```

# SMART Question 1  
**Did Medicare patients have, on average, at least 15% longer hospital stays than Private Insurance patients in Niagara County (2012)?**

```{r Q1, echo = FALSE, message = FALSE, warning = FALSE}



cat("Step 1: Loading the dataset\n")
data <- read.csv("Hospital_Inpatient_Discharges.csv")
cat("Dataset loaded successfully with", nrow(data), "rows and", ncol(data), "columns.\n\n")

# -----------------------------------------------------------
cat("Step 2: Handling missing values\n")
clean_data <- data
if (any(is.na(data))) {
  cat("Missing values detected!\n")
  total_missing <- sum(is.na(data))
  total_cells <- nrow(data) * ncol(data)
  if (total_missing / total_cells < 0.05) {
    clean_data <- na.omit(data)
    cat("Rows with missing values removed (less than 5% missing).\n")
  } else {
    for (col in names(clean_data)) {
      if (is.numeric(clean_data[[col]])) {
        clean_data[[col]][is.na(clean_data[[col]])] <- median(clean_data[[col]], na.rm = TRUE)
      }
    }
    cat("Missing values in numeric columns replaced with median values.\n")
  }
} else {
  cat("No missing values detected. Dataset is clean.\n")
}

cat("Removing duplicate rows if any...\n")
clean_data <- clean_data %>% distinct()
cat("Duplicates removed. Data ready for analysis.\n\n")

# -----------------------------------------------------------
cat("Step 3: Categorizing insurance types and converting stay length to numeric\n")
df <- data %>%
  mutate(
    Insurance_Group = case_when(
      Payment.Typology.1 == "Medicare" |
        Payment.Typology.2 == "Medicare" |
        Payment.Typology.3 == "Medicare" ~ "Medicare",
      Payment.Typology.1 %in% c("Blue Cross/Blue Shield", "Private Health Insurance") |
        Payment.Typology.2 %in% c("Blue Cross/Blue Shield", "Private Health Insurance") |
        Payment.Typology.3 %in% c("Blue Cross/Blue Shield", "Private Health Insurance") ~ "Private Insurance",
      TRUE ~ NA_character_
    ),
    Length.of.Stay = as.numeric(Length.of.Stay)
  ) %>%
  filter(!is.na(Insurance_Group), !is.na(Length.of.Stay)) %>%
  group_by(Insurance_Group) %>%
  filter(between(
    Length.of.Stay,
    quantile(Length.of.Stay, 0.25, na.rm = TRUE) - 1.5 * IQR(Length.of.Stay, na.rm = TRUE),
    quantile(Length.of.Stay, 0.75, na.rm = TRUE) + 1.5 * IQR(Length.of.Stay, na.rm = TRUE)
  )) %>%
  ungroup()

cat("Insurance groups categorized and outliers removed.\n\n")

# -----------------------------------------------------------
cat("Step 4: Computing descriptive statistics\n")
group_stats <- df %>%
  group_by(Insurance_Group) %>%
  summarise(
    mean_LOS = mean(Length.of.Stay, na.rm = TRUE),
    sd_LOS = sd(Length.of.Stay, na.rm = TRUE),
    count = n()
  )
print(group_stats)
cat("\nDescriptive statistics calculated for each insurance group.\n\n")

# -----------------------------------------------------------
cat("Step 5: Computing ratio of means (Medicare vs. Private)\n")
ratio <- group_stats$mean_LOS[group_stats$Insurance_Group == "Medicare"] /
  group_stats$mean_LOS[group_stats$Insurance_Group == "Private Insurance"]
cat("Mean Medicare LOS is", round(ratio, 2), "times that of Private Insurance.\n\n")

# -----------------------------------------------------------
cat("Step 6: Running one-sided t-test\n")
t_result <- t.test(
  Length.of.Stay ~ Insurance_Group,
  data = df,
  alternative = "greater",
  na.action = na.omit
)
print(t_result)
cat("T-test completed.\n\n")

# -----------------------------------------------------------
cat("Step 7: Interpreting results\n")
p_val <- t_result$p.value
medicare_mean <- group_stats$mean_LOS[group_stats$Insurance_Group == "Medicare"]
private_mean <- group_stats$mean_LOS[group_stats$Insurance_Group == "Private Insurance"]
diff_ratio <- (medicare_mean / private_mean)
percent_diff <- (diff_ratio - 1) * 100

if (!is.na(p_val) && !is.na(diff_ratio)) {
  if (p_val < 0.05 & diff_ratio >= 1.15) {
    cat("Medicare patients stayed significantly longer on average.\n",
        "The mean length of stay for Medicare patients (", round(medicare_mean, 2),
        " days) was about ", round(percent_diff, 1),
        "% longer than for Private Insurance patients (", round(private_mean, 2),
        " days), with p = ", signif(p_val, 4),
        ". This difference is statistically significant (p < 0.05).\n", sep = "")
  } else {
    cat("No statistically significant difference of at least 15% in average hospital stay ",
        "between Medicare and Private Insurance patients (p = ", signif(p_val, 4), ").\n", sep = "")
  }
} 

# -----------------------------------------------------------
cat("Step 8: Visualizing average hospital stays by insurance type\n")
ggplot(group_stats, aes(x = Insurance_Group, y = mean_LOS, fill = Insurance_Group)) +
  geom_bar(stat = "identity", width = 0.6, color = "black") +
  geom_text(aes(label = round(mean_LOS, 2)), vjust = -0.4, size = 4) +
  labs(
    title = "Average Hospital Length of Stay by Insurance Type (Niagara County, 2012)",
    x = "Insurance Type",
    y = "Mean Length of Stay (Days)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```
 
 
# SMART Question 2:
**Can we analyze 2012 hospital records to quantify whether Medicaid patients’ total hospital costs were 10% higher than patients with Private Insurance? **

```{r q2, message=FALSE, warning=FALSE}

# 1️⃣ Select relevant columns
columns_to_keep <- c("Facility.ID", "Age.Group", "Gender", "Length.of.Stay",
  "Payment.Typology.1", "Payment.Typology.2", "Payment.Typology.3",
  "Total.Charges", "Total.Costs")

hospital_final <- clean_data[, columns_to_keep]


# 2️⃣ Filter Medicaid and Private insurance types
insurance_pattern <- "blue cross|blue shield|bcbs|private insurance|medicaid"
hospital_filtered <- hospital_final %>%
  filter(
    str_detect(tolower(Payment.Typology.1), insurance_pattern) |
    str_detect(tolower(Payment.Typology.2), insurance_pattern) |
    str_detect(tolower(Payment.Typology.3), insurance_pattern)
  )


# 3️⃣ Create insurance group labels
values_to = "Insurance.Type"

hospital_filtered <- hospital_filtered %>%
  mutate(
    Insurance_Use = if_else(
      if_any(starts_with("Payment.Typology"), ~ str_detect(str_to_lower(.), "medicaid")),
      "Medicaid", "Private Insurance"
    )
  )

# 4️⃣ Remove outliers using ezids::outlierKD2()

cat("Removing outliers for Total.Costs:\n")
hospital_no_outliers <- outlierKD2(hospital_filtered, Total.Costs, rm = TRUE, qqplt = TRUE)

cat("Removing outliers for Total.Charges:\n")
hospital_no_outliers <- outlierKD2(hospital_no_outliers, Total.Charges, rm = TRUE, qqplt = TRUE)


# --- Statistical test (t-test) ---
cat("Apply T-test on Average Total Hospital Costs by Insurance Type:")
t_result <- t.test(`Total.Costs` ~ Insurance_Use, data = hospital_no_outliers)
print(t_result)




# --- Bar plot: Mean Total Costs by Insurance Type ---
cost_means <- hospital_no_outliers %>%
  group_by(Insurance_Use) %>%
  summarise(mean_cost = mean(`Total.Costs`, na.rm = TRUE))


ggplot(cost_means, aes(x = Insurance_Use, y = mean_cost, fill = Insurance_Use)) +
  geom_bar(stat = "identity", width = 0.5) +
  geom_text(aes(label = round(mean_cost, 0)), vjust = -0.5, size = 4) +
  labs(
    title = "Average Total Hospital Costs by Insurance Type",
    x = "Insurance Type",
    y = "Mean Total Costs ($)"
  ) +
  scale_fill_manual(values = c("Medicaid" = "#FF9999", "Private Insurance" = "#99CCFF")) +
  theme_minimal() +
  theme(legend.position = "none")  

# Print result

means <- t_result$estimate
different_percent <- (means[1] - means[2]) / means[2] * 100
ci <- t_result$conf.int

if(different_percent>0){
  cat(" Medicaid patients' total hospital costs are approximately", round(abs(different_percent), 2), "% higher than Private Insurance.\n")
}else{
   cat("Medicaid patients' total hospital costs are approximately", round(abs(different_percent), 2), "% lower than Private Insurance.\n")
  cat("Medicaid patients had lower mean total hospital costs ($", round(means[1],2), 
    ") than Private Insurance patients ($", round(means[2],2), "), approximately ", 
    round(abs(different_percent),2), "% difference. ",
    "The difference was statistically significant (t = ", round(t_result$statistic,2),
    ", df = ", round(t_result$parameter,1), ", p < 0.001).\n", sep = "")
}

cat("Extra part: Here we check the total_charge based on insurance types:")


# --- t-test for Total Charges ---
cat("Apply T-test on Average Total Hospital charge by Insurance Type:")
t_charge_result <- t.test(`Total.Charges` ~ Insurance_Use, data = hospital_no_outliers)
print(t_charge_result)

# --- Bar plot: Mean Total Charges by Insurance Type ---
cost_means <- hospital_no_outliers %>%
  group_by(Insurance_Use) %>%
  summarise(mean_cost = mean(`Total.Charges`, na.rm = TRUE))


ggplot(cost_means, aes(x = Insurance_Use, y = mean_cost, fill = Insurance_Use)) +
  geom_bar(stat = "identity", width = 0.5) +
  geom_text(aes(label = round(mean_cost, 0)), vjust = -0.5, size = 4) +
  labs(
    title = "Average Total Hospital Charges by Insurance Type",
    x = "Insurance Type",
    y = "Mean Total Charges ($)"
  ) +
  scale_fill_manual(values = c("Medicaid" = "#FF9999", "Private Insurance" = "#99CCFF")) +
  theme_minimal() +
  theme(legend.position = "none")  

# Print result

means <- t_charge_result$estimate
different_percent <- (means[1] - means[2]) / means[2] * 100
ci <- t_charge_result$conf.int

if(different_percent>0){
  cat(" Medicaid patients' total hospital charges are approximately", round(abs(different_percent), 2), "% higher than Private Insurance.\n")
}else{
   cat("Medicaid patients' total hospital charges are approximately", round(abs(different_percent), 2), "% lower than Private Insurance.\n")
  cat("Medicaid patients had lower mean total hospital charges ($", round(means[1],2), 
    ") than Private Insurance patients ($", round(means[2],2), "), approximately ", 
    round(abs(different_percent),2), "% difference. ",
    "The difference was statistically significant (t = ", round(t_charge_result$statistic,2),
    ", df = ", round(t_charge_result$parameter,1), ", p < 0.001).\n", sep = "")
}

```
# SMART Question 3  
**Did elderly patients (70+) stay 25% longer than patients aged 30–49 in 2012?**

```{r q3, echo=FALSE, include=TRUE, results='asis', message=FALSE, warning=FALSE}
# Use cleaned dataset
data_df <- clean_data
# Inspect columns
table(data_df$Discharge.Year, useNA="ifany")
table(data_df$Age.Group, useNA="ifany")
# Create age buckets
data_df <- data_df %>%
mutate(age_bucket = case_when(
  grepl("70", Age.Group, ignore.case = TRUE) ~ "70+",
  grepl("30", Age.Group, ignore.case = TRUE) & grepl("49", Age.Group, ignore.case = TRUE) ~ "30-49",
  TRUE ~ NA_character_
))
cat("Age buckets created. Here are the counts:\n")
print(table(data_df$age_bucket, useNA="ifany"))
# Clean Length of Stay
data_df$Length.of.Stay <- as.numeric(data_df$Length.of.Stay)
cat("Length.of.Stay cleaned. Sample values:\n")
print(head(data_df$Length.of.Stay, 10))
# Subsets
g70 <- data_df %>% filter(age_bucket == "70+") %>% pull(Length.of.Stay)
g30 <- data_df %>% filter(age_bucket == "30-49") %>% pull(Length.of.Stay)
cat("Number of 70+ patients:", length(g70), "\n")
cat("Number of 30-49 patients:", length(g30), "\n")
# Descriptive stats
summary_stats <- data_df %>%
group_by(age_bucket) %>%
summarise(
n = n(),
mean = mean(Length.of.Stay, na.rm=TRUE),
median = median(Length.of.Stay, na.rm=TRUE),
sd = sd(Length.of.Stay, na.rm=TRUE)
)
summary_stats
cat("Descriptive statistics by age bucket:\n")
print(summary_stats)
#percent difference
mean70 <- mean(g70, na.rm=TRUE)
mean30 <- mean(g30, na.rm=TRUE)
percent_longer <- (mean70 - mean30) / mean30 * 100
cat("\nPercent longer stay for 70+ vs 30–49:", round(percent_longer,2), "%\n")
cat("Count of observations:\n")
cat("70+ group:", sum(!is.na(g70)), "\n")
cat("30-49 group:", sum(!is.na(g30)), "\n")
# Check data availability
n70 <- sum(!is.na(g70))
n30 <- sum(!is.na(g30))

cat("Number of observations:\n")
cat("70+ group:", n70, "\n")
cat("30–49 group:", n30, "\n")

# Run t-test only if both groups have at least 2 valid observations
if (n70 > 1 & n30 > 1) {
  tres <- t.test(g70, g30, var.equal = FALSE)
  print(tres)
} else {
  cat("\n❌ Not enough data for Welch t-test. Skipping this step.\n")
}
# SMART Question 3: Do elderly patients (70+) stay 25% longer than patients aged 30–49?

# 1. Load dataset
df <- clean_data

# 2. Clean and inspect Age.Group
df$Age.Group <- trimws(df$Age.Group)
cat("Unique Age Groups:\n")
print(unique(df$Age.Group))

# 3. Flexible filter for the two target groups
df_age <- df %>%
  filter(
    (grepl("30", Age.Group, ignore.case = TRUE) & grepl("49", Age.Group, ignore.case = TRUE)) |
    grepl("70", Age.Group, ignore.case = TRUE)
  ) %>%
  droplevels()

# 4. Clean Length.of.Stay
df_age$Length.of.Stay <- suppressWarnings(as.numeric(df_age$Length.of.Stay))
df_age <- df_age %>% filter(!is.na(Length.of.Stay))

# 5. Check resulting groups
cat("\nGroups after filtering:\n")
print(table(df_age$Age.Group, useNA="ifany"))

# Proceed only if two groups with valid data
if (length(unique(df_age$Age.Group)) == 2 &&
    all(sapply(split(df_age$Length.of.Stay, df_age$Age.Group), function(x) length(x) > 1))) {
  
  # Group means
  age_means <- df_age %>%
    group_by(Age.Group) %>%
    summarise(mean_LOS = mean(Length.of.Stay, na.rm = TRUE))
  print(age_means)
  
  # Calculate ratio
  ratio_age <- age_means$mean_LOS[grepl("70", age_means$Age.Group)] /
               age_means$mean_LOS[grepl("30", age_means$Age.Group)]
  cat("\nElderly / 30–49 ratio =", round(ratio_age, 2), "\n")
  
  # One-sided t-test (elderly > younger)
  cat("\nWelch Two Sample t-test:\n")
  print(t.test(Length.of.Stay ~ Age.Group, data = df_age, alternative = "greater"))
  
  # Boxplot
  ggplot(df_age, aes(x = Age.Group, y = Length.of.Stay, fill = Age.Group)) +
    geom_boxplot() +
    theme_minimal() +
    labs(title = "Length of Stay by Age Group (Niagara)",
         x = "Age Group", y = "Length of Stay (days)")
  
} else {
  cat("\n❌ Cannot run t-test. Check data integrity.\n")
}


cat("If p-value < 0.05\n",
    "Since the p-value is less than 0.05, we reject the null hypothesis.\n",
    "This means there is a statistically significant difference between the two group means.\n",
    "In other words, the observed difference is unlikely to have occurred by random chance.\n\n")

cat("If p-value ≥ 0.05\n",
    "Since the p-value is greater than 0.05, we fail to reject the null hypothesis.\n",
    "This indicates there is no statistically significant difference between the group means.\n",
    "The difference observed could be due to random variation.\n")
```

# SMART Question 4
**Did emergency admissions result in 30% longer hospital stays than elective ones in 2012?**

```{r q4}


library(tidyverse)
library(janitor)
library(ggplot2)
library(car)
library(knitr)


# --- Title & Introduction ---------------------------------------------------

cat("
Purpose:
Analyze the difference in Length of Stay (LOS) between Emergency and Elective admissions
using SPARCS 2012 Niagara County data.

Why it matters:
Hospital efficiency and patient-flow optimization rely on understanding LOS variation.
")


# --- Data Preparation -------------------------------------------------------

df <- read_csv("Hospital_Inpatient_Discharges.csv")
dfq <- df %>%
clean_names() %>%
filter(hospital_county == "Niagara") %>%
distinct() %>%
transmute(
los = as.numeric(length_of_stay),
admission_type = factor(type_of_admission),
age_group = factor(age_group),
gender = factor(gender)
) %>%
drop_na(los, admission_type) %>%
  filter(admission_type %in% c("Emergency", "Elective"))

kable(head(dfq))
summary(dfq)



# --- EDA Visuals ------------------------------------------------------------



# Violin + Box + CI

ggplot(dfq, aes(admission_type, los, fill = admission_type)) +
geom_violin(trim = FALSE, alpha = 0.5) +
geom_boxplot(width = 0.2, alpha = 0.7, color = "black") +
stat_summary(fun.data = mean_cl_normal, geom = "pointrange", color = "red") +
labs(title = "Length of Stay by Admission Type (Niagara County)",
x = "Admission Type", y = "Length of Stay (Days)") +
theme_minimal()


# Histogram

ggplot(dfq, aes(los, fill = admission_type)) +
geom_histogram(position = "identity", alpha = 0.5, bins = 30) +
facet_wrap(~ admission_type, scales = "free_y") +
theme_minimal() +
labs(title = "Distribution of LOS by Admission Type",
x = "Length of Stay", y = "Frequency")


# Mean LOS by Age Group

dfq %>%
group_by(age_group, admission_type) %>%
summarise(mean_los = mean(los, na.rm = TRUE)) %>%
ggplot(aes(age_group, mean_los, fill = admission_type)) +
geom_col(position = "dodge") +
labs(title = "Mean LOS by Age Group and Admission Type",
x = "Age Group", y = "Average LOS (Days)") +
theme_minimal()


# Mean LOS by Gender

dfq %>%
group_by(gender, admission_type) %>%
summarise(mean_los = mean(los, na.rm = TRUE)) %>%
ggplot(aes(gender, mean_los, fill = admission_type)) +
geom_col(position = "dodge") +
labs(title = "Mean LOS by Gender and Admission Type",
x = "Gender", y = "Average LOS (Days)") +
theme_minimal()



# --- Assumption Checks ------------------------------------------------------



ggplot(dfq, aes(sample = los)) +
stat_qq() + stat_qq_line(color = "blue") +
facet_wrap(~ admission_type) +
theme_minimal() +
labs(title = "QQ Plot of LOS by Admission Type")

leveneTest(los ~ admission_type, data = dfq)



# --- Welch’s t-test (emergency vs elective only) ----------------------------



dfq_test <- dfq %>%
  filter(admission_type %in% c("Emergency", "Elective")) %>%
  droplevels()

unique(dfq_test$admission_type)

t_res <- t.test(los ~ admission_type, data = dfq_test, var.equal = FALSE)
t_res





#### --- Percent Difference (to test the 30% from question) -------------------

mean_values <- dfq_test %>%
  group_by(admission_type) %>%
  summarise(mean_los = mean(los, na.rm = TRUE), .groups = "drop")

mean_values

mean_emergency <- mean_values$mean_los[mean_values$admission_type == "Emergency"]
mean_elective  <- mean_values$mean_los[mean_values$admission_type == "Elective"]

# Percent difference: (E - L) / L * 100
perc_diff <- (mean_emergency - mean_elective) / mean_elective * 100

# Direction-aware message
if (perc_diff >= 0) {
  cat("Emergency stays are", round(perc_diff, 2), "% LONGER than Elective stays.\n")
} else {
  cat("Emergency stays are", round(abs(perc_diff), 2), "% SHORTER than Elective stays.\n")
}





# --- Ratio check for the '30% longer' hypothesis ----------------------------



ratio_obs <- mean_emergency / mean_elective
cat("Observed mean ratio (Emergency/Elective):", round(ratio_obs, 3), "\n")

if (ratio_obs >= 1.3) {
  message("Emergency stays meet or exceed the 30% longer hypothesis.")
} else {
  message("Emergency stays do NOT meet the 30% longer hypothesis.")
}



# --- Effect Size + Log Sensitivity -----------------------------------------




if ("effectsize" %in% rownames(installed.packages())) {
library(effectsize)
cohens_d(los ~ admission_type, data = dfq_test)
} else {
los_emergency <- dfq_test$los[dfq_test$admission_type == "Emergency"]
los_elective  <- dfq_test$los[dfq_test$admission_type == "Elective"]
mean_diff <- mean(los_emergency) - mean(los_elective)
pooled_sd <- sqrt(((sd(los_emergency)^2 + sd(los_elective)^2) / 2))
cohen_d <- mean_diff / pooled_sd
cat("Manual Cohen’s d =", round(cohen_d, 3), "\n")
}

dfq_test <- dfq_test %>% mutate(log_los = log1p(los))
t.test(log_los ~ admission_type, data = dfq_test)



# --- Regression (Age & Gender) ------------------------------------



lm_los <- lm(los ~ admission_type + age_group + gender, data = dfq)
summary(lm_los)



# --- final takeway --------------------------------------------------------------


cat("
• Emergency patients stay significantly SHORTER than elective patients.
• Observed mean ratio (Emergency/Elective) ≈ ", round(mean_emergency/mean_elective, 3), ".
• Welch t-test: p-value from t_res = ", signif(t_res$p.value, 3), ".
• Effect size is small-to-moderate (report the actual Cohen's d below).
• Implication: elective (often surgical) cases drive longer LOS; plan capacity accordingly.
")


```