---
title: "Final Project: Examining the Relationship Between Demographics, Admission Types, and Clinical Outcomes in New York Inpatients (Niagara County)"
author: "Team 2: Simintaj Salehpour, Madeline Bumpus, Nazish Atta, Divya Mistry"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    number_sections: true
    df_print: paged
    code_folding: hide
    self_contained: true
---

<style>
body {
  background-color: #f4f8fb;
  color: #222;
  font-family: "Lato", "Helvetica Neue", sans-serif;
}
h1, h2, h3 {
  color: #004d80;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)
library(kableExtra)
library(skimr)
library(naniar)
```

#Introduction

We examined differences in hospital length of stay (LOS) across insurance types to explore how coverage influences healthcare utilization. Our focus was on Medicare and Private Insurance, with additional attention to Medicaid, combining personal perspectives with data-driven insights to understand disparities in care and resource use.

#Data Description

We used the SPARCS Hospital Inpatient Discharges dataset for Niagara County, 2024, containing ~6,913 patient discharge records. It includes demographics, diagnoses, procedures, insurance type, admission type, and clinical outcomes. This publicly available dataset enables analysis of LOS and other healthcare metrics at the county level.

```{r Load Dataset, include=T, results='markup'}
data <- read.csv("/Users/divyamistry/Downloads/Hospital_Inpatient_Discharges.csv")
names(data) <- make.names(names(data))  # clean column names
cat("Dataset dimensions: Rows:", nrow(data), "Columns:", ncol(data), "\n")
```

```{r Missing values, include=T, results='markup'}
#Summary of missing values
missing_summary <- data %>%
summarise(across(everything(), ~sum(is.na(.)))) %>%
pivot_longer(everything(), names_to = "variable", values_to = "n_missing") %>%
arrange(desc(n_missing))
missing_summary
#Visualize
gg_miss_var(data) + labs(title = "Variables vs Number of Missing Values")
```

```{r Remove Duplicates, include=T, results='markup'}
#Remove Duplicates
dup_count <- sum(duplicated(data))
cat("Number of duplicate rows:", dup_count, "\n")
clean_data <- data %>% distinct()
```

## SMART Question: Readmission Risk Prediction
## Question: Which combination of patient age group, severity of illness, and payer type predicts the lowest hospital readmission risk?

```{r Q3, message = FALSE, warning = FALSE, include=TRUE}

library(dplyr)
library(ggplot2)
library(caret)
library(pROC)

set.seed(123)
#1. Create binary outcome: 1 = died in hospital, 0 = survived

clean_data$Mortality <- ifelse(clean_data$Patient.Disposition == "Expired", 1L, 0L)

#2. Filter out missing values for predictors

q3_data <- clean_data %>%
filter(
!is.na(Age.Group),
!is.na(APR.Severity.of.Illness.Description),
!is.na(Payment.Typology.1),
!is.na(Mortality)
) %>%
select(
Mortality,
Age.Group,
APR.Severity.of.Illness.Description,
Payment.Typology.1
)

#3. Set variable types

q3_data$MortalityClass <- factor(q3_data$Mortality,
levels = c(0,1),
labels = c("Survived","Died"))
q3_data$Age.Group <- droplevels(as.factor(q3_data$Age.Group))
q3_data$APR.Severity.of.Illness.Description <- droplevels(as.factor(q3_data$APR.Severity.of.Illness.Description))
q3_data$Payment.Typology.1 <- droplevels(as.factor(q3_data$Payment.Typology.1))

#4. Train/Test split

index_q3 <- createDataPartition(q3_data$MortalityClass, p = 0.7, list = FALSE)
train_q3 <- q3_data[index_q3, ]
test_q3 <- q3_data[-index_q3, ]

#5. Logistic regression model

logit_q3 <- glm(
MortalityClass ~ Age.Group +
APR.Severity.of.Illness.Description +
Payment.Typology.1,
data = train_q3,
family = binomial
)

summary(logit_q3)

#6. Predicted probabilities on TEST set

pred_prob_q3 <- predict(logit_q3, newdata = test_q3, type = "response")

#7. ROC curve and AUC

roc_q3 <- roc(test_q3$Mortality, pred_prob_q3)
auc_q3 <- auc(roc_q3)

plot(roc_q3,
col = "#0066CC",
lwd = 2,
main = paste("ROC Curve for In-Hospital Mortality (AUC =",
round(auc_q3, 3), ")"))

#8. Classify "high-risk" patients at cutoff = 0.10

cutoff <- 0.10
pred_class_num <- ifelse(pred_prob_q3 >= cutoff, 1L, 0L)
pred_class_fac <- factor(pred_class_num,
levels = c(0, 1),
labels = c("Survived", "Died"))

#Confusion matrix

conf_mat_q3 <- confusionMatrix(
pred_class_fac,
test_q3$MortalityClass,
positive = "Died"
)

conf_mat_q3

#Extract key metrics

accuracy <- conf_mat_q3$overall["Accuracy"]
sensitivity <- conf_mat_q3$byClass["Sensitivity"]
specificity <- conf_mat_q3$byClass["Specificity"]
ppv <- conf_mat_q3$byClass["Pos Pred Value"]
npv <- conf_mat_q3$byClass["Neg Pred Value"]

#9. Visualization: Mortality by Severity and Payer

ggplot(train_q3,
aes(x = APR.Severity.of.Illness.Description,
fill = MortalityClass)) +
geom_bar(position = "fill", color = "black") +
scale_fill_manual(values = c("Survived" = "#FF9933", "Died" = "#0066CC")) +
facet_wrap(~ Payment.Typology.1) +
labs(
title = "Proportion of In-Hospital Deaths by Severity and Payer Type",
x = "Severity of Illness",
y = "Proportion within Group",
fill = "Outcome"
) +
theme_bw() +
theme(
plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
axis.title = element_text(size = 12),
axis.text = element_text(size = 10),
legend.title = element_text(size = 11),
legend.text = element_text(size = 10)
)

#10. Mortality rate summary for each predictor

mortality_summary <- bind_rows(
q3_data %>%
group_by(APR.Severity.of.Illness.Description) %>%
summarise(mortality_rate = mean(Mortality)) %>%
mutate(Factor = "Severity of Illness", Category = APR.Severity.of.Illness.Description),

q3_data %>%
group_by(Age.Group) %>%
summarise(mortality_rate = mean(Mortality)) %>%
mutate(Factor = "Age Group", Category = Age.Group),

q3_data %>%
group_by(Payment.Typology.1) %>%
summarise(mortality_rate = mean(Mortality)) %>%
mutate(Factor = "Payer Type", Category = Payment.Typology.1)
)
mortality_summary$Category <- as.factor(mortality_summary$Category)

ggplot(mortality_summary,
aes(x = Category, y = mortality_rate * 100, fill = Factor)) +
geom_col(show.legend = FALSE) +
facet_wrap(~ Factor, scales = "free_x") +
labs(
title = "Comparison of In-Hospital Mortality Rates Across Predictors",
x = NULL,
y = "Mortality Rate (%)"
) +
theme_bw() +
theme(
strip.text = element_text(face = "bold", size = 12),
axis.text.x = element_text(angle = 30, hjust = 1)
)

#11. Summary & Conclusion

cat("\n--- SMART Question: Mortality Prediction Summary ---\n")
cat("AUC:", round(auc_q3, 3), "\n")
cat("Accuracy:", round(accuracy, 3), "\n")
cat("Sensitivity:", round(sensitivity, 3), "\n")
cat("Specificity:", round(specificity, 3), "\n")
cat("PPV:", round(ppv, 3), "\n")
cat("NPV:", round(npv, 3), "\n")
cat("\n--- Conclusion ---\n")
cat("The logistic regression model indicates that Age Group, Severity of Illness, and Payer Type all influence in-hospital mortality risk.\n")
cat("Patients with higher severity and certain payer types (e.g., Medicare) show higher mortality risk.\n")
cat("You can use this model to identify high-risk patients and guide clinical decision-making.\n")
```