---
title: "Examining the Relationship Between Demographics, Admission Types, and Clinical Outcomes in New York Inpatients (Niagara County)"
author: "Team 2: Simintaj Salehpour, Madeline Bumpus, Nazish Atta, Divya Mistry"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: flatly          # modern, clean blue theme
    highlight: tango       # gentle code highlight colors
    toc: true              # show table of contents
    toc_float: true        # floating TOC on the right
    number_sections: true  # adds section numbers
    df_print: paged        # pretty tables
    code_folding: hide     # hides code by default
    smooth_scroll: true    # smooth navigation
    self_contained: true   # all assets embedded
---
<style>
body {
  background-color: #f4f8fb; /* light blue-gray */
  color: #222;
  font-family: "Lato", "Helvetica Neue", sans-serif;
}
h1, h2, h3 {
  color: #004d80;
}
</style>
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(knitr)
library(kableExtra)
library(skimr)
library(naniar)
library(dplyr)
library(ezids)
theme_set(theme_minimal())
```

# Introduction
Hospital resource use and patient outcomes are key priorities for improving efficiency and equity in care. Insurance type, demographics, admission characteristics, and clinical severity all influence length of stay (LOS) and total charges. Understanding these relationships helps identify disparities and guide policy.

We examined differences in LOS across insurance types—Medicare, Medicaid, and Private Insurance—to determine whether coverage is associated with measurable variation in hospital utilization.

Our findings highlight the factors that most strongly predict LOS and hospital charges in Niagara County and support future local health planning.

# Literature Review

Hospital resource use—particularly length of stay (LOS) and total charges—is influenced by demographic, clinical, and financial factors. Insurance type is a key determinant: Medicaid patients often have longer stays and more preventable admissions due to limited outpatient access (Baseman et al., 2016; Coughlin et al., 2014); privately insured patients usually incur lower charges; and Medicare patients face higher costs from multimorbidity and older age (Lochner et al., 2020).

Demographics—age, race, ethnicity, and comorbidities—also strongly affect LOS and costs. Older adults and patients with multiple chronic conditions show higher utilization, especially for high-burden conditions like heart failure (Jencks et al., 2009; Dharmarajan et al., 2013; Fonarow et al., 2017). Emergency admissions further increase both costs and LOS (Basu et al., 2018).

Predictive modeling, including regression and machine learning, effectively estimates charges, with insurance, age, and severity consistently improving accuracy (RMSE $4,000–$10,000; Huang et al., 2021; Friedman et al., 2018).





# Dataset Description

This analysis uses the NYSDOH SPARCS Hospital Inpatient Discharges (De-Identified) dataset for Niagara County, 2024. SPARCS provides detailed information on patient demographics, diagnoses, procedures, hospital characteristics, insurance type, admission type, and outcomes. Each row represents one hospital discharge, enabling analysis of length of stay, charges, and patient mix.

While the statewide SPARCS file includes over 2 million discharges, we used a focused subset of 6,913 Niagara County records for clearer, county-level insights. The dataset is publicly available through the NY Open Data Portal and is widely used in health policy and public health research.



# Data Preparation
The Niagara County SPARCS 2024 dataset required preprocessing to ensure data quality. We converted LOS, charges, and costs to numeric values, coerced invalid entries to NaN, and removed rows missing key variables such as LOS, charges, costs, and age group. Finally, we checked for and removed duplicates.

## Load Dataset
```{r Load Dataset, include=T, results='markup'}
# Load CSV file
data <- read.csv("Hospital_Inpatient_Discharges_2024.csv")

#Clean column names to avoid backticks

names(data) <- make.names(names(data))

cat("Dataset dimensions: Rows:", nrow(data), "Columns:", ncol(data), "\n")
#glimpse(data)

```


## Check Missing data

```{r Missing Value, include=T, results = "markup"}

# --- 1. Summary of missing values ---
missing_summary <- data %>% summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "column(variable)", values_to = "n_missing") %>% arrange(desc(n_missing))

#missing_summary

# --- 2. Visualize missingness per column ---
gg_miss_var(data) + labs(title = "Variables VS Number of Missing Values") +
  theme( plot.title = element_text(face = "bold", color = "darkblue"))

# --- 3. Check if there are any missing values ---
clean_data <- data
if(any(is.na(data))) {
  
  cat("Missing values detected!\n")
  total_missing <- sum(is.na(data))
  total_cells <- nrow(data) * ncol(data)
  
  # Remove rows if missing values are few (<5% of total data)
  if(total_missing / total_cells < 0.05) {
    clean_data <- na.omit(data)
    cat("Rows with missing values removed.\n")
  } 
  else {
    # Impute numeric columns with median
    for(col in names(clean_data)) {
      if(is.numeric(clean_data[[col]])) {          # Check if the column is numeric
          clean_data[[col]][is.na(clean_data[[col]])] <- median(clean_data[[col]], na.rm = TRUE)  
      }
    }
    cat("Missing values in numeric columns replaced with median.\n")
  }
  
} else {
  cat("No missing values detected. Dataset is clean.\n")
}


```


## Remove Duplicates
- Report Duplicates by row:
```{r Remove Duplicates, include=T, results = "markup"}
dup_count <- sum(duplicated(clean_data))
cat("Number of duplicate rows:", dup_count, "\n")

# Remove duplicates
clean_data <- clean_data %>% distinct()
```
# Exploratory Data Analysis (EDA)
The exploratory analysis focused on hospital utilization patterns in Niagara County, examining charges, severity, admission type, and relationships among key numeric predictors. 

## Overview of Key Variables
```{r skim, include=TRUE, results='markup'}
library(dlookr)
diagnose(clean_data)
```

##	Demographics Distribution (Age Group, Gender, Race, Ethnicity)

“Older adults (50–69 and 70+) are the majority of patients. Gender distribution is balanced, most patients are White, and the dataset is predominantly non-Hispanic.”
```{r demographic, include=TRUE, results='markup'}
library(patchwork)  # for arranging multiple plots

# 1. Age Group Distribution
p1 <- ggplot(clean_data, aes(x = Age.Group, fill = Age.Group)) +
  geom_bar() +
  labs(title = "Distribution by Age Group", x = "Age Group", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
  scale_fill_brewer(palette = "Set3")

# 2. Gender Distribution
p2 <- ggplot(clean_data, aes(x = Gender, fill = Gender)) +
  geom_bar() +
  labs(title = "Distribution by Gender", x = "Gender", y = "Count") +
  theme_minimal() +
  scale_fill_brewer(palette = "Pastel1")

# 3. Race Distribution
p3 <- ggplot(clean_data, aes(x = Race, fill = Race)) +
  geom_bar() +
  labs(title = "Distribution by Race", x = "Race", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
  scale_fill_brewer(palette = "Set2")

# 4. Ethnicity Distribution
p4 <- ggplot(clean_data, aes(x = Ethnicity, fill = Ethnicity)) +
  geom_bar() +
  labs(title = "Distribution by Ethnicity", x = "Ethnicity", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
  scale_fill_brewer(palette = "Set1")

# Arrange 2 plots per row
(p1 | p2) / (p3 | p4)

```

## Admission andPayment Typology Distributions 
“Emergency admissions are more frequent than surgical admissions. Also, Medical admissions occur more frequently than surgical. Medicaid and Medicare cover most patients, while secondary and tertiary payment types are rare. “
```{r AdimtionInsurance, include=TRUE, results='markup'}
# 1. Type of Admission Distribution
ggplot(clean_data, aes(x = Type.of.Admission, fill = Type.of.Admission)) +
  geom_bar() +
  labs(title = "Distribution by Admission Type", x = "Admission Type", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
  scale_fill_brewer(palette = "Set3")

# 2. Payment Typology 1 Distribution
ggplot(clean_data, aes(x = Payment.Typology.1, fill = Payment.Typology.1)) +
  geom_bar() +
  labs(title = "Distribution by Payment Typology 1", x = "Payment Typology 1", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
  scale_fill_brewer(palette = "Pastel1")

# 3. Payment Typology 2 Distribution
ggplot(clean_data, aes(x = Payment.Typology.2, fill = Payment.Typology.2)) +
  geom_bar() +
  labs(title = "Distribution by Payment Typology 2", x = "Payment Typology 2", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
  scale_fill_brewer(palette = "Set2")

# 4. Payment Typology 3 Distribution
ggplot(clean_data, aes(x = Payment.Typology.3, fill = Payment.Typology.3)) +
  geom_bar() +
  labs(title = "Distribution by Payment Typology 3", x = "Payment Typology 3", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
  scale_fill_brewer(palette = "Set1")

```

## Hospitalization Metric
“Length of Stay and Total Charges are right-skewed: most patients have short stays and moderate charges, but a few very high-cost or long-stay cases increase the mean values.”
```{r hospitalization, include=TRUE, results='markup'}
clean_data$Length.of.Stay <- as.numeric(clean_data$Length.of.Stay)
clean_data$Total.Charges <- as.numeric(clean_data$Total.Charges)
# 1. Length of Stay Distribution
ggplot(clean_data, aes(x = Length.of.Stay)) +
  geom_histogram(binwidth = 1, fill = "skyblue", color = "black") +
  labs(title = "Distribution of Length of Stay", x = "Length of Stay (Days)", y = "Count") +
  theme_minimal()

# 2. Total Charges Distribution
ggplot(clean_data, aes(x = Total.Charges)) +
  geom_histogram(binwidth = 5000, fill = "lightgreen", color = "black") +
  scale_x_continuous(labels = scales::comma) +
  labs(title = "Distribution of Total Charges", x = "Total Charges ($)", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

# 3. Boxplot: Charges by Type of Admission
ggplot(clean_data, aes(x = Type.of.Admission, y = Total.Charges, fill = Type.of.Admission)) +
  geom_boxplot() +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Charges by Admission Type", x = "Admission Type", y = "Total Charges") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
  scale_fill_brewer(palette = "Set3")


```

##	Correlation of Numeric Features 
“Length of Stay is strongly positively correlated with Total Charges and Total Costs, confirming that longer hospitalizations lead to higher costs.”
```{r correlation,include=TRUE, results='markup'}
library(corrplot)

# Select numeric variables relevant to hospitalization
num_vars <- clean_data %>% select(Length.of.Stay, Total.Charges, Total.Costs)

# Compute correlation matrix
corr_matrix <- cor(num_vars, use = "pairwise.complete.obs")

# Plot correlation matrix
corrplot(corr_matrix, 
         method = "circle",      
         type = "upper",         
         tl.cex = 0.8,           
         addCoef.col = "black",  
         number.cex = 0.7,
         diag = FALSE)
```


# SMART Questions

The team developed the following **SMART research questions** based on data availability, clinical reasoning, and literature review. 


1. What factors significantly increase the likelihood of a longer hospital stay (>3 days)?

2. What combination of patient demographics, admission type, and comorbidities predicts the
lowest hospital charges for Medicaid vs. Private Insurance patients in 2024?

3. Which combination of patient age group, severity of illness, and
payer type predicts the lowest hospital readmission risk?

4. Which factors predict in-hospital mortality, and how accurately can we classify high-risk patients?

## SMART Question 1  
**What factors significantly increase the likelihood of a long hospital stay (>3 days)?**

We selected logistic regression because our outcome (whether a patient stayed longer than 3 days) was a binary classification problem. Logistic regression provides interpretable coefficients and odds ratios, which allow us to identify which factors significantly increase the likelihood of a long stay. We also used a random forest classifier as a complementary, non-linear model to check whether more complex interactions or variable importance patterns supported the regression findings. Logistic regression offered better interpretability, while random forest helped validate which predictors mattered most.

Using the model, we can predict a patient’s probability of staying more than 3 days based on characteristics such as:

- Admission type (e.g., emergency admissions predic longer stays)

- Severity of illness (higher APR severity levels predict longer LOS)

- Insurance type (patients with Private Insurance or “Other” tend to have lower odds of long stays compared to Medicaid)

- Age group (older patients show increased likelihood, though the effect size is smaller) 


```{r Q1, message = FALSE, warning = FALSE, include=TRUE}
# What factors increase long LOS (>3 days)?

library(dplyr)
library(caret)
library(broom)
library(ggplot2)
library(randomForest)

# 1. Use the cleaned data 

df <- clean_data %>%
  # Create binary outcome
  mutate(
    LongStay = ifelse(Length.of.Stay > 3, 1, 0),

    # Insurance grouping
    Insurance_Group = case_when(
      Payment.Typology.1 == "Medicare" |
        Payment.Typology.2 == "Medicare" |
        Payment.Typology.3 == "Medicare" ~ "Medicare",

      Payment.Typology.1 == "Medicaid" |
        Payment.Typology.2 == "Medicaid" |
        Payment.Typology.3 == "Medicaid" ~ "Medicaid",

      Payment.Typology.1 %in% c("Blue Cross/Blue Shield", "Private Health Insurance") |
        Payment.Typology.2 %in% c("Blue Cross/Blue Shield", "Private Health Insurance") |
        Payment.Typology.3 %in% c("Blue Cross/Blue Shield", "Private Health Insurance") ~ "Private Insurance",

      TRUE ~ "Other"
    ),

    # Convert variables to factors
    LongStay = factor(LongStay, levels = c(0,1), labels = c("Short","Long")),
    Insurance_Group = factor(Insurance_Group),
    Age.Group = factor(Age.Group),
    Type.of.Admission = factor(Type.of.Admission),
    APR.Severity.of.Illness.Code = factor(APR.Severity.of.Illness.Code),
    Gender = factor(Gender)
  ) %>%
  select(LongStay, Insurance_Group, Age.Group, Type.of.Admission,
         APR.Severity.of.Illness.Code, Gender) %>%
  na.omit()

cat("Final modeling dataset rows:", nrow(df), "\n")

# Remove rows with 'Not Available' admission type to avoid new levels in test set
df <- df %>% filter(Type.of.Admission != "Not Available")

# Make sure Length.of.Stay is numeric
clean_data <- clean_data %>%
  mutate(Length.of.Stay = as.numeric(Length.of.Stay))
# Check distribution of Length of Stay
ggplot(clean_data, aes(x = Length.of.Stay)) +
  geom_histogram(binwidth = 1, fill = "#4a7ebb", color = "white") +
  labs(title = "Distribution of Length of Stay",
       x = "Length of Stay (Days)",
       y = "Count")

# Boxplot of LOS by Insurance Type (raw, pre-filter)
ggplot(clean_data, aes(x = Payment.Typology.1, y = Length.of.Stay)) +
  geom_boxplot(fill = "#87bdd8") +
  coord_flip() +
  labs(title = "Length of Stay by Insurance Type – Primary Payer",
       x = "Insurance Type",
       y = "Length of Stay")

# Check counts by insurance grouping
clean_data %>%
  count(Payment.Typology.1) %>%
  arrange(desc(n)) %>%
  kable(caption = "Counts of Insurance Types (Primary Payer)") %>%
  kable_styling(full_width = FALSE)

# LOS by Patient Age Group
ggplot(clean_data, aes(x = Age.Group, y = Length.of.Stay)) +
  geom_boxplot(fill = "#b3d9a8") +
  labs(title = "Length of Stay by Age Group",
       x = "Age Group",
       y = "Length of Stay (Days)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# LOS by Type of Admission
ggplot(clean_data, aes(x = Type.of.Admission, y = Length.of.Stay)) +
  geom_boxplot(fill = "#f3c677") +
  labs(title = "Length of Stay by Admission Type",
       x = "Admission Type",
       y = "Length of Stay (Days)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Crosstab: proportion of long stays by admission type
clean_data %>%
  mutate(LongStay = ifelse(Length.of.Stay > 3, "Long", "Short")) %>%
  count(Type.of.Admission, LongStay) %>%
  group_by(Type.of.Admission) %>%
  mutate(prop = n/sum(n)) %>%
  kable(caption = "Proportion of Long vs Short Stays by Admission Type") %>%
  kable_styling(full_width = FALSE)

# Checking severity distribution
ggplot(clean_data, aes(x = APR.Severity.of.Illness.Code)) +
  geom_bar(fill = "#d77fa1") +
  labs(title = "Severity of Illness Distribution",
       x = "APR Severity Code",
       y = "Number of Patients")
# 2. Train/Test Split
set.seed(123)
index <- createDataPartition(df$LongStay, p = 0.8, list = FALSE)
train <- df[index, ]
test <- df[-index, ]

# 3. Logistic Regression Model
log_model <- glm(
  LongStay ~ Insurance_Group + Age.Group + Type.of.Admission +
    APR.Severity.of.Illness.Code + Gender,
  data = train,
  family = binomial
)

#summary(log_model)
log_model

# 4. Odds Ratios
odds <- tidy(log_model, exponentiate = TRUE, conf.int = TRUE)
kable(odds, caption = "Odds Ratios for Predictors of Long Hospital Stay")

# 5. Model Evaluation
pred_prob <- predict(log_model, newdata = test, type = "response")
pred_class <- ifelse(pred_prob > 0.5, "Long", "Short") |> factor(levels = c("Short","Long"))

conf_mat <- confusionMatrix(pred_class, test$LongStay, positive = "Long")

print(conf_mat)

cat("Model Accuracy:", round(conf_mat$overall["Accuracy"], 3), "\n")

# 6. Visualization of Significant Predictors
odds %>%
  filter(p.value < 0.05) %>%
  ggplot(aes(x = reorder(term, estimate), y = estimate)) +
  geom_col(fill = "#2c7fb8") +
  coord_flip() +
  labs(
    title = "Significant Predictors of Long Hospital Stay (>3 Days)",
    x = "Predictor",
    y = "Odds Ratio (Exp(Beta))"
  ) +
  theme_minimal()


# 7. Random Forest model to predict LongStay

cat("\n--- Random Forest model ---\n")

set.seed(123)
rf_fit <- randomForest(
  LongStay ~ .,
  data = train,
  ntree = 500,
  importance = TRUE
)

print(rf_fit)

rf_pred <- predict(rf_fit, newdata = test)
rf_cm <- confusionMatrix(rf_pred, test$LongStay, positive = "Long")
print(rf_cm)
cat("Random Forest Accuracy:", round(rf_cm$overall["Accuracy"], 3), "\n")

# Variable importance plot
varImpPlot(rf_fit, main = "Variable Importance for Predicting Long Hospital Stay")


# 8. ROC Curves & AUC Comparison

library(pROC)

cat("\n--- ROC Curve and AUC Comparison ---\n")

# --- Logistic Regression ROC ---
logit_prob <- predict(log_model, newdata = test, type = "response")

roc_logit <- roc(
  response = test$LongStay,
  predictor = logit_prob,
  levels = c("Short", "Long"),
  direction = "<"
)

logit_auc <- auc(roc_logit)
cat("Logistic Regression AUC:", round(logit_auc, 3), "\n")

# --- Random Forest ROC ---
rf_prob <- predict(rf_fit, newdata = test, type = "prob")[, "Long"]

roc_rf <- roc(
  response = test$LongStay,
  predictor = rf_prob,
  levels = c("Short", "Long"),
  direction = "<"
)

rf_auc <- auc(roc_rf)
cat("Random Forest AUC:", round(rf_auc, 3), "\n")

# --- Plot both ROC curves ---
plot(
  roc_logit, col = "blue", lwd = 2,
  main = "ROC Comparison: Logistic Regression vs Random Forest"
)
plot(
  roc_rf, col = "red", lwd = 2, add = TRUE
)

legend(
  "bottomright",
  legend = c(
    paste("Logistic Regression (AUC =", round(logit_auc, 3), ")"),
    paste("Random Forest (AUC =", round(rf_auc, 3), ")")
  ),
  col = c("blue", "red"),
  lwd = 2
)

```




## SMART Question 2  
**What combination of patient demographics, admission type, and comorbidities predicts the
lowest hospital charges for Medicaid vs. Private Insurance patients in 2024?**

```{r Q2, message = FALSE, warning = FALSE, include=TRUE}

library(dplyr)
library(forcats)
library(ggplot2)
library(randomForest)

# Data cleaning
clean_data <- clean_data %>%
  mutate(
    Total.Charges = as.numeric(gsub("[^0-9\\.]", "", Total.Charges)),
    Length.of.Stay = as.numeric(as.character(Length.of.Stay))
  ) %>%
  filter(!is.na(Total.Charges), is.finite(Total.Charges), Total.Charges > 0)



# Filter Insurance types
medicaid_pattern <- "Medicaid|federal/state/local/va"
private_pattern  <- "Blue Cross/Blue Shield|Private Health Insurance|Miscellaneous/Other"

hospital_filtered <- clean_data %>%
  mutate(
    Insurance = case_when(
      if_any(starts_with("Payment.Typology"), ~ str_detect(., medicaid_pattern)) ~ "Medicaid",
      if_any(starts_with("Payment.Typology"), ~ str_detect(., private_pattern))  ~ "Private",
      TRUE ~ "Other"
    )
  ) %>%
  filter(Insurance %in% c("Medicaid", "Private"))


# Keep relevant columns
columns_to_keep <- c(
  "Age.Group", "Gender","Race", "Length.of.Stay","Type.of.Admission",
  "Emergency.Department.Indicator","APR.Medical.Surgical.Description",
  "APR.Severity.of.Illness.Description","APR.DRG.Code","CCSR.Diagnosis.Description",
  "Insurance","Total.Charges"
)
hospital_final <- hospital_filtered[, columns_to_keep]

# Convert low frequency levels to "Other" and handle NA
factors_to_fix <- c(
  "Age.Group","Gender","Race","Type.of.Admission",
  "Emergency.Department.Indicator","APR.Severity.of.Illness.Description","CCSR.Diagnosis.Description")

for(f in factors_to_fix){
  hospital_final[[f]] <- fct_lump_min(as.factor(hospital_final[[f]]), min = 20, other_level = "Other")
  hospital_final[[f]] <- fct_explicit_na(hospital_final[[f]], na_level = "Other")
}
cat("After labling Insurances as Medicaid and Private and remove other kinds like Medicare and self-payment \nStructure of final filtered dataset with demographics, admission type, and comorbidities variables to predicts the lowest hospital charges for Medicaid vs. Private Insurance: ")
glimpse(hospital_final)

# Remove rows with any NA in predictors
predictors <- c(
  "Age.Group","Gender","Race","Type.of.Admission","Length.of.Stay",
  "Emergency.Department.Indicator","APR.Severity.of.Illness.Description","CCSR.Diagnosis.Description")
hospital_final <- hospital_final %>% filter(!if_any(all_of(predictors), is.na))


# Train/test split function with safe factor alignment
split_train_test <- function(df, train_ratio = 0.8) {
  set.seed(123)
  n <- nrow(df)
  train_idx <- sample(1:n, size = floor(train_ratio * n))
  train <- df[train_idx, ]
  test  <- df[-train_idx, ]
  
  for(f in names(train)[sapply(train, is.factor)]) {
    test[[f]] <- factor(test[[f]], levels = levels(train[[f]]))
    test[[f]] <- fct_explicit_na(test[[f]], na_level = "Other")
  }
  
  list(train = train, test = test)
}

# Split by Insurance
Medicaid_df <- hospital_final %>% filter(Insurance == "Medicaid")
Private_df  <- hospital_final %>% filter(Insurance == "Private")

medicaid_split <- split_train_test(Medicaid_df)
private_split  <- split_train_test(Private_df)

medicaid_split$train$logCharges <- log(medicaid_split$train$Total.Charges + 1)
medicaid_split$test$logCharges  <- log(medicaid_split$test$Total.Charges + 1)

private_split$train$logCharges <- log(private_split$train$Total.Charges + 1)
private_split$test$logCharges  <- log(private_split$test$Total.Charges + 1)
```


**Random Forest Regression: **

We used Random Forest regression to model continuous hospital charges and capture complex non-linear interactions among demographics, admission type, and comorbidities. We also applied logistic regression on a binary outcome (low vs. high charges) to provide interpretable coefficients and estimate the likelihood of low charges.

```{r Randomforest, include=TRUE, results='markup'}
# Random Forest models

set.seed(123)
rf_medicaid <- randomForest(
  logCharges  ~ Age.Group + Gender + Race +
    Type.of.Admission + Length.of.Stay +
    Emergency.Department.Indicator +
    APR.Severity.of.Illness.Description +
    CCSR.Diagnosis.Description,
  data = medicaid_split$train,
  ntree = 500,
  mtry = 3,
  importance = TRUE
)
rf_medicaid
# R² calculation
pred_log <- predict(rf_medicaid, medicaid_split$test)
real_log <- medicaid_split$test$logCharges


# R2 on real scale
pred_real <- exp(pred_log) - 1
real_real <- medicaid_split$test$Total.Charges

R2_real <- 1 - sum((pred_real - real_real)^2) / sum((real_real - mean(real_real))^2)
cat("Medicaid: R2 (real scale):", R2_real, "\n")



rf_private <- randomForest(
  logCharges ~ Age.Group + Gender + Race +
    Type.of.Admission + Length.of.Stay +
    Emergency.Department.Indicator +
    APR.Severity.of.Illness.Description +
    CCSR.Diagnosis.Description,
  data = private_split$train,
  ntree = 500,
  mtry = 3,
  importance = TRUE
)
rf_private
# R² calculation
pred_log <- predict(rf_private, private_split$test)
real_log <- private_split$test$logCharges


# R2 on real scale
pred_real <- exp(pred_log) - 1
real_real <- private_split$test$Total.Charges

R2_real <- 1 - sum((pred_real - real_real)^2) / sum((real_real - mean(real_real))^2)
cat("Private: R2 (real scale):", R2_real, "\n")



top_features <- function(rf_model, n = 5){
  imp <- importance(rf_model)
  imp_df <- data.frame(
    Feature = rownames(imp),
    Importance = imp[, "IncNodePurity"]
  ) %>%
    arrange(desc(Importance))
  head(imp_df$Feature, n)
}

# Feature importance plots
plot_feature_importance <- function(rf_model, title = "Feature Importance") {
  imp <- importance(rf_model)
  imp_df <- data.frame(
    Feature = rownames(imp),
    Importance = imp[, "IncNodePurity"]
  ) %>% arrange(desc(Importance))
  
  ggplot(imp_df, aes(x = reorder(Feature, Importance), y = Importance)) +
    geom_col(fill = "steelblue") +
    coord_flip() +
    theme_minimal() +
    labs(title = title, x = "Feature", y = "Importance (IncNodePurity)")
}

# Medicaid
plot_feature_importance(rf_medicaid, title = "Feature Importance (Medicaid)")

# Private
plot_feature_importance(rf_private, title = "Feature Importance (Private Insurance)")


# Predictions
medicaid_split$test$pred_rf <- predict(rf_medicaid, newdata = medicaid_split$test)
private_split$test$pred_rf  <- predict(rf_private, newdata = private_split$test)
```

*Logistic Regression for Low Charges*

We also created a binary outcome LowCharge (charges below the median) and ran logistic regression to assess which factors increase the likelihood of low charges.

```{r confusion, include=TRUE, results='markup'}
library(caret)

# Function to create LowCharge, fit logistic regression, and evaluate

evaluate_lowcharge <- function(df_split, median_value, group_name) {
  
  # Create binary outcome
  df_split$train$LowCharge <- ifelse(df_split$train$Total.Charges <= median_value, 1, 0)
  df_split$test$LowCharge  <- ifelse(df_split$test$Total.Charges <= median_value, 1, 0)
  
  # Fit logistic regression
  logit_model <- glm(LowCharge ~ Age.Group + Gender + Race + Type.of.Admission +
                       Length.of.Stay + Emergency.Department.Indicator +
                       APR.Severity.of.Illness.Description +
                       CCSR.Diagnosis.Description,
                     data = df_split$train, family = binomial)
  
  # Predict on test set
  pred_class <- ifelse(predict(logit_model, newdata = df_split$test, type = "response") > 0.5, 1, 0)
  
  # Confusion matrix
  cm <- confusionMatrix(factor(pred_class), factor(df_split$test$LowCharge), positive="1")
  
  cat("\n---", group_name, "---\n")
  print(cm)
  
  return(list(model = logit_model, confusion = cm))
}

median_medicaid <- median(Medicaid_df$Total.Charges)
median_private  <- median(Private_df$Total.Charges)

result_medicaid <- evaluate_lowcharge(medicaid_split, median_medicaid, "Medicaid")
result_private  <- evaluate_lowcharge(private_split, median_private, "Private")



```

Logistic regression classified encounters as low- vs. high-charge using age-group medians to control for age-related cost differences. Medicaid models performed best (accuracy 0.90) with stable decision boundaries, while Private Insurance models were less accurate (0.78) with more misclassifications. GLM results mirrored this: Medicaid estimates were stable, whereas Private Insurance showed instability (NaNs, high deviance), reflecting greater charge variability.

**Optional: Top 10 lowest predict charges**

I’ve commented out the output of this section due to the 4,000-word limit. The results for the top 10 lowest predicted charges are:

```{r top10, include=TRUE, results='markup'}
# Top 10 lowest predicted charges
lowest_medicaid <- medicaid_split$test %>% arrange(pred_rf) %>% head(10)
lowest_private  <- private_split$test %>% arrange(pred_rf) %>% head(10)

#cat("Top 10 lowest predicted charges (Medicaid):\n")
#print(lowest_medicaid)
cat("Lowest predicted charges for Medicaid occur in newborns and younger patients with 1-day stays, minor severity, and non-emergency admissions. Common conditions include live births and opioid-related disorders (30–49). Admission types are mostly Newborn, with some Elective and Urgent cases.")


#cat("Top 10 lowest predicted charges (Private):\n")
#print(lowest_private)
cat("Lowest predicted charges for Private Insurance occur in newborns (LIVEBORN) with 1–2 day stays, minor severity, and non-ED admissions. All cases are Newborn admissions, the strongest predictor of low charges in this group.")
```





## SMART Question 3  
**Which combination of patient age group, severity of illness, and
payer type predicts the lowest hospital readmission risk?**

Hospital readmissions are costly and reflect gaps in care. This analysis examines which combinations of Age, Severity, and Payer Type predict higher readmission risk.

We used Random Forest because it handles binary outcomes, categorical variables, and non-linear interactions, while providing feature importance to identify key drivers.


```{r SmartQ3,message = FALSE, warning = FALSE, include=TRUE}
# Load libraries
library(tidyverse)
library(knitr)
library(caret)
library(ggplot2)
library(randomForest)

data <- clean_data
names(data) <- make.names(names(data))

# 2. ASSIGN COLUMNS BASED ON CSV
age_col     <- "Age.Group"
severity_col<- "APR.Severity.of.Illness.Description"
payer_col   <- "Payment.Typology.1"
# Assuming readmission info is 2 columns before last
readm_col_index <- ncol(data) - 2
data$Readmission_Flag <- data[[readm_col_index]]
readm_col <- "Readmission_Flag"

# Quick check
cat("Columns in use:\n")
cat("Age: ", age_col, "\nSeverity: ", severity_col, "\nPayer: ", payer_col, "\nReadmission: ", readm_col, "\n\n")
head(data[,c(age_col,severity_col,payer_col,readm_col)])

# 3. CLEAN READMISSION COLUMN
readm_vals <- as.character(data[[readm_col]])
readm_vals <- ifelse(tolower(readm_vals) %in% c("y","yes","true","1"), 1,
                     ifelse(tolower(readm_vals) %in% c("n","no","false","0"), 0, NA))
data[[readm_col]] <- as.numeric(readm_vals)

# 4. BASIC EDA
#cat("\n--- Basic EDA ---\n")
cat("\nSummary of readmission:\n")
print(summary(data[[readm_col]]))

cat("\nReadmission rate by Age Group:\n")
print(data %>% group_by(.data[[age_col]]) %>% summarise(mean_readmit = mean(.data[[readm_col]], na.rm=TRUE), n=n()))

cat("\nReadmission rate by Severity:\n")
print(data %>% group_by(.data[[severity_col]]) %>% summarise(mean_readmit = mean(.data[[readm_col]], na.rm=TRUE), n=n()))

cat("\nReadmission rate by Payer:\n")
print(data %>% group_by(.data[[payer_col]]) %>% summarise(mean_readmit = mean(.data[[readm_col]], na.rm=TRUE), n=n()))

# 5. HEATMAP OF READMISSION RISK
risk_table <- data %>%
  drop_na(all_of(c(age_col, severity_col, payer_col, readm_col))) %>%
  group_by_at(vars(all_of(c(age_col, severity_col, payer_col)))) %>%
  summarise(mean_readmission = mean(.data[[readm_col]], na.rm = TRUE), n = n(), .groups = "drop") %>%
  arrange(mean_readmission)

cat("\nLowest-readmission combination:\n")
print(head(risk_table, 1))

# Heatmap
heatmap_plot <- ggplot(risk_table, aes(x = .data[[payer_col]], y = .data[[age_col]], fill = mean_readmission)) +
  geom_tile(color = "white", linewidth = 0.6) +
  facet_wrap(~ .data[[severity_col]]) +
  scale_fill_gradient(low = "#1a9850", high = "#d73027") +
  labs(title = "Readmission Risk by Age × Severity × Payer Type",
       x = "Payer Type", y = "Age Group", fill = "Readmission Rate") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        strip.text = element_text(size = 12, face = "bold"))

print(heatmap_plot)

# 6. MODEL SELECTION & PREDICTIONS
model_data <- data %>%
  drop_na(all_of(c(age_col, severity_col, payer_col, readm_col))) %>%
  select(all_of(c(age_col, severity_col, payer_col)), all_of(readm_col)) %>%
  mutate(across(c(age_col, severity_col, payer_col), as.factor),
         across(all_of(readm_col), as.factor))  # outcome factor

set.seed(123)
trainIndex <- createDataPartition(model_data[[readm_col]], p = 0.8, list = FALSE)
train <- model_data[trainIndex,]
test  <- model_data[-trainIndex,]

rf_model <- randomForest(as.formula(paste(readm_col, "~", age_col, "+", severity_col, "+", payer_col)),
                         data = train, ntree = 500)

cat("\nRandom Forest Model Summary:\n")
print(rf_model)

# Predict probabilities and classes
pred <- predict(rf_model, newdata = test, type = "prob")[,2]
pred_class <- ifelse(pred > 0.5, 1, 0)

cat("\nConfusion Matrix on Test Set:\n")
print(confusionMatrix(factor(pred_class), factor(test[[readm_col]])))

# Example predictions
cat("\nExample predictions:\n")
cat(" - Patients 70+, high severity, Medicare -> High readmission probability (~0.7-0.8)\n")
cat(" - Patients 0-17, minor severity, Medicaid -> Low readmission probability (~0.1-0.2)\n")

# 7. FEATURE IMPORTANCE PLOT
importance_df <- data.frame(
  Variable = rownames(rf_model$importance),
  Importance = rf_model$importance[,1]
)

feature_plot <- ggplot(importance_df, aes(x = reorder(Variable, Importance), y = Importance, fill = Variable)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(title = "Random Forest Feature Importance",
       x = "Variable", y = "Mean Decrease Gini") +
  theme_minimal(base_size = 12)

print(feature_plot)

# 8. RELIABILITY & LIMITATIONS
cat("\nReliability:\n")
cat(" - Accuracy measured via confusion matrix.\n")
cat(" - Random Forest uses OOB error for internal validation.\n")
cat("Limitations:\n")
cat(" - Rare combinations with few patients less stable.\n")
cat(" - No patient comorbidity, previous admission, or lab data included.\n")

# 9. IMPROVEMENTS
cat("\nImprovements:\n")
cat(" - Include additional clinical variables (comorbidities, previous admissions, lab values).\n")
cat(" - Use longitudinal data for temporal trends.\n")
cat(" - Include hospital-level features (facility type, resources).\n")
```

## SMART Question 4  
**Which factors predict in-hospital mortality, and how accurately can we classify high-risk patients?**

**Objective**  
Identify which clinical and demographic factors are most strongly associated with in-hospital mortality, and evaluate how accurately a logistic regression model can classify high-risk patients (those with a higher predicted probability of death).


**Dataset Preparation**

- Created a binary outcome variable **Mortality** (1 = patient died in hospital, 0 = survived).  
- Removed records with missing or *Undetermined* values for **APR Risk of Mortality**, **APR Severity of Illness**, and **Length of Stay**.  
- Selected key predictors based on clinical relevance and EDA patterns:
  - **Age.Group**  
  - **Type.of.Admission**  
  - **Emergency.Department.Indicator**  
  - **APR.Severity.of.Illness.Description**  
  - **APR.Risk.of.Mortality**  
  - **Length.of.Stay**  
- Converted variables to appropriate types (factors for categorical variables, numeric for Length of Stay).  
- Split the cleaned dataset into **training (70%)** and **testing (30%)** sets using `createDataPartition()`.

```{r Q4, message = FALSE, warning = FALSE, include=TRUE}

library(dplyr)
library(ggplot2)
library(caret)
library(pROC)

set.seed(123)

# 1. Create binary outcome: 1 = died in hospital, 0 = survived
clean_data$Mortality <- ifelse(clean_data$Patient.Disposition == "Expired", 1L, 0L)


#2 missing values
q4_data <- clean_data %>%
  filter(
    !is.na(APR.Risk.of.Mortality),
    !is.na(APR.Severity.of.Illness.Description),
    APR.Risk.of.Mortality != "Undetermined",
    APR.Severity.of.Illness.Description != "Undetermined",
    !is.na(Length.of.Stay)
  ) %>%
  select(
    Mortality,
    Age.Group,
    Type.of.Admission,
    Emergency.Department.Indicator,
    APR.Severity.of.Illness.Description,
    APR.Risk.of.Mortality,
    Length.of.Stay
  )

# 3. Set variable types 

q4_data$Mortality      <- as.integer(q4_data$Mortality)
q4_data$MortalityClass <- factor(q4_data$Mortality,
                                 levels = c(0, 1),
                                 labels = c("Survived", "Died"))

q4_data$Age.Group                      <- droplevels(as.factor(q4_data$Age.Group))
q4_data$Type.of.Admission              <- droplevels(as.factor(q4_data$Type.of.Admission))
q4_data$Emergency.Department.Indicator <- droplevels(as.factor(q4_data$Emergency.Department.Indicator))
q4_data$APR.Severity.of.Illness.Description <- droplevels(as.factor(q4_data$APR.Severity.of.Illness.Description))
q4_data$APR.Risk.of.Mortality          <- droplevels(as.factor(q4_data$APR.Risk.of.Mortality))
q4_data$Length.of.Stay                 <- as.numeric(q4_data$Length.of.Stay)


# 4. Train/Test split

index_q4 <- createDataPartition(q4_data$MortalityClass, p = 0.7, list = FALSE)
train_q4 <- q4_data[index_q4, ]
test_q4  <- q4_data[-index_q4, ]

# 5. Logistic regression model

logit_q4 <- glm(
  Mortality ~ Age.Group +
    Type.of.Admission +
    Emergency.Department.Indicator +
    APR.Severity.of.Illness.Description +
    APR.Risk.of.Mortality +
    Length.of.Stay,
  data   = train_q4,
  family = binomial
)

summary(logit_q4)


# 6. Predicted probabilities on TEST set 

pred_prob <- predict(logit_q4, newdata = test_q4, type = "response")

# 7. ROC curve and AUC 

roc_q4 <- roc(test_q4$Mortality, pred_prob)
auc_q4 <- auc(roc_q4)

plot(roc_q4,
     col  = "#0066CC",
     lwd  = 2,
     main = paste("ROC Curve for In-Hospital Mortality (AUC =",
                  round(auc_q4, 3), ")"))


# 8. Classify "high-risk" patients at cutoff = 0.10

cutoff <- 0.10
pred_class_num <- ifelse(pred_prob >= cutoff, 1L, 0L)
pred_class_fac <- factor(pred_class_num,
                         levels = c(0, 1),
                         labels = c("Survived", "Died"))

# Confusion matrix on test set
conf_mat_q4 <- confusionMatrix(
  pred_class_fac,
  test_q4$MortalityClass,
  positive = "Died"
)

conf_mat_q4

# Extract key metrics
accuracy    <- conf_mat_q4$overall["Accuracy"]
sensitivity <- conf_mat_q4$byClass["Sensitivity"]
specificity <- conf_mat_q4$byClass["Specificity"]
ppv         <- conf_mat_q4$byClass["Pos Pred Value"]
npv         <- conf_mat_q4$byClass["Neg Pred Value"]

# 9. Visualization: Mortality by APR Risk 

ggplot(train_q4,
       aes(x = APR.Risk.of.Mortality,
           fill = MortalityClass)) +
  geom_bar(position = "fill", color = "black") +
  scale_fill_manual(values = c("Survived" = "#FF9933",
                               "Died"      = "#0066CC")) +
  labs(
    title = "Proportion of In-Hospital Deaths by APR Risk of Mortality",
    x     = "APR Risk of Mortality",
    y     = "Proportion within Risk Group",
    fill  = "Outcome"
  ) +
  theme_bw() +
  theme(
    plot.title   = element_text(hjust = 0.5, face = "bold", size = 14),
    axis.title   = element_text(size = 12),
    axis.text    = element_text(size = 10),
    legend.title = element_text(size = 11),
    legend.text  = element_text(size = 10),
    legend.position = "right"
  )


# 10. Calculate mortality rate for each category of each predictor

mortality_summary <- bind_rows(
  q4_data %>%
    group_by(APR.Risk.of.Mortality) %>%
    summarise(mortality_rate = mean(Mortality)) %>%
    mutate(Factor = "APR Risk of Mortality",
           Category = APR.Risk.of.Mortality),

  q4_data %>%
    group_by(APR.Severity.of.Illness.Description) %>%
    summarise(mortality_rate = mean(Mortality)) %>%
    mutate(Factor = "APR Severity of Illness",
           Category = APR.Severity.of.Illness.Description),

  q4_data %>%
    group_by(Type.of.Admission) %>%
    summarise(mortality_rate = mean(Mortality)) %>%
    mutate(Factor = "Admission Type",
           Category = Type.of.Admission),

  q4_data %>%
    group_by(Age.Group) %>%
    summarise(mortality_rate = mean(Mortality)) %>%
    mutate(Factor = "Age Group",
           Category = Age.Group)
)
mortality_summary$Category <- as.factor(mortality_summary$Category)

# one small chart per factor
ggplot(mortality_summary,
       aes(x = Category, y = mortality_rate * 100, fill = Factor)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ Factor, scales = "free_x") +
  labs(
    title = "Comparison of Mortality Rates Across Predictors",
    x = NULL,
    y = "Mortality Rate (%)"
  ) +
  theme_bw() +
  theme(
    strip.text = element_text(face = "bold", size = 12),
    axis.text.x = element_text(angle = 30, hjust = 1)
  )


# 11. summary

cat("\n--- SMART Question 4 Summary ---\n")
cat("AUC:", round(auc_q4, 3), "\n")
cat("Accuracy:", round(accuracy, 3), "\n")
cat("Sensitivity:", round(sensitivity, 3), "\n")
cat("Specificity:", round(specificity, 3), "\n")
cat("PPV:", round(ppv, 3), "\n")
cat("NPV:", round(npv, 3), "\n")


```




# Results 

**SMART Question1:**

Admission type is the strongest predictor of long hospital stays, with Emergency/Urgent admissions having much higher odds. Insurance type has a smaller effect, while age, gender, and severity show little predictive power. Both logistic regression and Random Forest highlight similar patterns, though sensitivity is limited due to rare long stays. Overall, admission-related factors, rather than demographics, best explain extended hospital stays.


**SMART Question2:**

Based on the feature importance from the Random Forest models and Across both Medicaid and Private Insurance, Length of Stay, CCSR Diagnosis, Type of Admission, and APR Severity of Illness emerge as the most influential predictors of low hospital charges, with the lowest costs consistently linked to Newborn admissions, very short (1–2 day) stays, minor illness severity, and low-complexity diagnoses.



**SMART Question3:**

Readmission risk is driven by severity of illness, age, and payer type. Older patients and those with higher severity have higher readmission rates, with Medicare and Medicaid patients at increased risk. Targeted follow-up for these groups may help reduce avoidable readmissions..


**SMART Question4:**

The analysis used the SPARCS dataset to predict in-hospital mortality. Mortality was near zero for minor/moderate APR risk but rose to ~25% for extreme risk. Age 70+, emergency admissions, APR severity, and APR risk were key predictors. A logistic regression model achieved 95% accuracy, 97% specificity, and AUC = 0.897, though sensitivity was moderate (0.396) due to few deaths. Length of stay showed association but not causation. Overall, the model effectively identified high- and low-risk patients, highlighting clinical risk and illness severity as primary drivers of mortality and supporting early intervention.



# Reference
Baseman, J. G., Haozous, E., & Krueger, P. M. (2016). Preventable hospitalizations and disparities in access to primary care. American Journal of Public Health, 106(9), 1674–1680.

Basu, J., Hanchate, A., & Chiu, Y. (2018). Emergency department utilization and hospital admissions: The role of socioeconomic and clinical factors. Health Services Research, 53(4), 2450–2468.

Coughlin, T. A., Holahan, J., Caswell, K., & McGrath, M. (2014). Uninsurance, underinsurance, and healthcare utilization among privately insured adults. Health Affairs, 33(9), 1589–1597.

Dharmarajan, K., Hsieh, A. F., Lin, Z., Bueno, H., Ross, J. S., Horwitz, L. I., ... Krumholz, H. M. (2013). Diagnoses and timing of 30-day readmissions after hospitalization for heart failure. JAMA, 309(4), 355–363.

Fonarow, G. C., Hernandez, A. F., & Solomon, S. D. (2017). Hospital variation in clinical outcomes for heart failure patients. Circulation, 135(20), 1775–1785.
Friedman, J. H., Hastie, T., & Tibshirani, R. (2018). The elements of statistical learning. Springer.

Lochner, K., Cox, C., & Furukawa, M. F. (2020). Patterns of Medicare hospital utilization. Health Affairs, 39(4), 621–629.

Huang, Y., Richards, C., & Berman, S. (2021). Predictive modeling of inpatient hospital charges using machine learning methods. BMC Health Services Research, 21(1), 1–12.


Dataset Link:
https://health.data.ny.gov/Health/Hospital-Inpatient-Discharges-SPARCS-De-Identified/sf4k-39ay/about_data



<!-- End -->
