---
title: "Final Project:Analyzing Demographics, Admission Types, Insurance Variation, and Critical Outcomes in Niagara County Inpatient Discharges (2024)"
author: "Team 2: Simintaj Salehpour,Madeline Bumpus, Nazish Atta, Divya Mistry"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: flatly          # modern, clean blue theme
    highlight: tango       # gentle code highlight colors
    toc: true              # show table of contents
    toc_float: true        # floating TOC on the right
    number_sections: true  # adds section numbers
    df_print: paged        # pretty tables
    code_folding: hide     # hides code by default
    smooth_scroll: true    # smooth navigation
    self_contained: true   # all assets embedded
---
<style>
body {
  background-color: #f4f8fb; /* light blue-gray */
  color: #222;
  font-family: "Lato", "Helvetica Neue", sans-serif;
}
h1, h2, h3 {
  color: #004d80;
}
</style>
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(knitr)
library(kableExtra)
library(skimr)
library(naniar)
library(dplyr)
library(ezids)
theme_set(theme_minimal())
```

# Introduction


Hospital utilization, cost differences, and patient outcomes are key for improving healthcare  quality, and analyzing how demographics, insurance type, comorbidities, and admission 
characteristics relate to length of stay, charges, and mortality provides valuable insights for planning and policy.
Our team chose to examine differences in **hospital length of stay (LOS)** across **insurance types** because it highlights an important intersection between healthcare access, cost, and patient outcomes. The relationship between insurance coverage and hospital utilization is central to understanding disparities in care delivery and efficiency within the U.S. healthcare system. 
We were particularly interested in comparing **Medicare**, which primarily serves older adults, and **Private Insurance**, which often provides broader access to care, to see how coverage type influences hospital resource use. Members of our team who have personally dealt with **Medicaid** and have seen firsthand how insurance status can impact the quality, timing, and length of medical care also find personal significance in this topic. We sought to better understand how insurance differences translate into quantifiable healthcare outcomes by combining our personal viewpoints with data-driven evidence through the analysis of actual hospital discharge data.
For the final project, our group analyzed Niagara County inpatient discharges (2024) using the SPARCS De-Identified dataset. Unlike the midterm, we have fully cleaned the dataset, completed exploratory analysis, and finalized the feature groups for predictive modeling.


# Data Description

## Datasets used

This analysis uses data from the **New York State Department of Health (NYSDOH) Statewide Planning and Research Cooperative System (SPARCS)**, specifically the *Hospital Inpatient Discharges (SPARCS De-Identified)* dataset for **Niagara County, 2024**.
The dataset includes detailed records of all hospital discharges statewide, with variables describing patient demographics, diagnoses, procedures, hospital facility characteristics, insurance/payment type, admission type, and clinical outcomes. Each row represents one patient discharge, allowing for analysis of trends in hospital length of stay, cost, and patient mix.
The full dataset contains over **2 million records** across New York State; for this project, a subset of approximately **6,913 discharges from Niagara County** was used to maintain manageability and statistical power. This dataset is publicly available through the **NY Open Data Portal** and is widely used in healthcare policy, public health, and data science research.




```{r Load Dataset, include=T, results='markup'}
# Load CSV file
data <- read.csv("Hospital_Inpatient_Discharges_2024.csv")

#Clean column names to avoid backticks

names(data) <- make.names(names(data))

cat("Dataset dimensions: Rows:", nrow(data), "Columns:", ncol(data), "\n")
#glimpse(data)

```


## Missing data
- Report missingness by variable:
We check if there is a missing value or not. For each column check the missing value.

```{r Missing Value, include=T, results = "markup"}

# --- 1. Summary of missing values ---
missing_summary <- data %>% summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "column(variable)", values_to = "n_missing") %>% arrange(desc(n_missing))

#missing_summary

# --- 2. Visualize missingness per column ---
gg_miss_var(data) + labs(title = "Variables VS Number of Missing Values") +
  theme( plot.title = element_text(face = "bold", color = "darkblue"))

# --- 3. Check if there are any missing values ---
clean_data <- data
if(any(is.na(data))) {
  
  cat("Missing values detected!\n")
  total_missing <- sum(is.na(data))
  total_cells <- nrow(data) * ncol(data)
  
  # Remove rows if missing values are few (<5% of total data)
  if(total_missing / total_cells < 0.05) {
    clean_data <- na.omit(data)
    cat("Rows with missing values removed.\n")
  } 
  else {
    # Impute numeric columns with median
    for(col in names(clean_data)) {
      if(is.numeric(clean_data[[col]])) {          # Check if the column is numeric
          clean_data[[col]][is.na(clean_data[[col]])] <- median(clean_data[[col]], na.rm = TRUE)  
      }
    }
    cat("Missing values in numeric columns replaced with median.\n")
  }
  
} else {
  cat("No missing values detected. Dataset is clean.\n")
}


```


## Remove Duplicates
- Report Duplicates by row:
 Check to remove duplicates:
```{r Remove Duplicates, include=T, results = "markup"}
dup_count <- sum(duplicated(clean_data))
cat("Number of duplicate rows:", dup_count, "\n")

# Remove duplicates
clean_data <- clean_data %>% distinct()
```
# SMART Questions

The team developed the following **SMART research questions** based on data availability, clinical reasoning, and literature review. These focus on measurable outcomes—length of stay and costs—to analyze healthcare disparities.


1. **SMART Question1**

2. **SMART Question2**

3. **SMART Question3**

4. **SMART Question4**



## SMART Question 1  
**Question1**

```{r Q1, message = FALSE, warning = FALSE, include=TRUE}



```


## SMART Question 2  
**What combination of patient demographics, admission type, and comorbidities predicts the lowest hospital charges for Medicaid vs. Private Insurance patients in 2024?**

```{r Q2, message = FALSE, warning = FALSE, include=TRUE}

clean_data <- clean_data %>%
  mutate(Total.Charges = as.numeric(gsub("[^0-9\\.]", "", Total.Charges))) %>%
  filter(!is.na(Total.Charges), is.finite(Total.Charges))

clean_data$Length.of.Stay<-as.numeric(as.character(clean_data$Length.of.Stay))


# Filter Medicaid and Private insurance types

medicaid_pattern <- "Medicaid|federal/state/local/va"
private_pattern  <- "Blue Cross/Blue Shield|Private Health Insurance|Miscellaneous/Other"

hospital_filtered <- clean_data %>%
  mutate(
    Insurance = case_when(
      if_any(starts_with("Payment.Typology"), ~ str_detect(., medicaid_pattern)) ~ "Medicaid",
      if_any(starts_with("Payment.Typology"), ~ str_detect(., private_pattern))  ~ "Private",
      TRUE ~ "Other"
    )
  ) %>%
  filter(Insurance %in% c("Medicaid", "Private"))

columns_to_keep <- c("Age.Group", "Gender","Race", "Length.of.Stay","Type.of.Admission","Emergency.Department.Indicator","APR.Medical.Surgical.Description","APR.Severity.of.Illness.Description","APR.DRG.Code","CCSR.Diagnosis.Description","Insurance",
  "Total.Charges")

hospital_final <- hospital_filtered[, columns_to_keep]


# #Build the model for each payer

# Ensure numeric variables are clean
hospital_final <- hospital_final %>%
  filter(Total.Charges > 0) %>%  # Gamma requires positive outcome
  mutate(across(c(Age.Group, Gender, Race, Type.of.Admission,
                   Length.of.Stay,Emergency.Department.Indicator,
                  APR.Severity.of.Illness.Description,
                  CCSR.Diagnosis.Description, Insurance),
                as.factor))
cat("Structure of final filtered dataset with demographics, admission type, and comorbidities variables to predicts the lowest hospital charges for Medicaid vs. Private Insurance ")
str(hospital_final)



Medicaid_df <- hospital_final %>% filter(Insurance == "Medicaid") %>% drop_na()
Private_df  <- hospital_final %>% filter(Insurance == "Private") %>% drop_na()

# -----------------------------
# Medicaid model (Gamma GLM)
# -----------------------------
model_medicaid <- glm(
  Total.Charges ~ Age.Group + Gender + Race + 
    Type.of.Admission + Length.of.Stay +
    Emergency.Department.Indicator +
    APR.Severity.of.Illness.Description +
    CCSR.Diagnosis.Description,
  family = Gamma(link = "log"),
  data = Medicaid_df
)

# Predictions (all positive)
Medicaid_df$pred <- predict(model_medicaid, type = "response")

# -----------------------------
# Private Insurance model (Gamma GLM)
# -----------------------------
model_private <- glm(
  Total.Charges ~ Age.Group + Gender + Race + 
    Type.of.Admission + Length.of.Stay +
    Emergency.Department.Indicator +
    APR.Severity.of.Illness.Description +
    CCSR.Diagnosis.Description,,
  family = Gamma(link = "log"),
  data = Private_df
)

# Predictions (all positive)
Private_df$pred <- predict(model_private, type = "response")

# -----------------------------
# Extract lowest predicted charges
# -----------------------------
lowest_medicaid <- Medicaid_df %>% arrange(pred) %>% head(10)
lowest_private  <- Private_df %>% arrange(pred) %>% head(10)

# View results
cat("Top 10 lowest predicted charges for Medicaid Insurance")
lowest_medicaid
cat("Top 10 lowest predicted charges for Private Insurance")
lowest_private





```



## SMART Question 3  
**Question3**

```{r Q3, message = FALSE, warning = FALSE, include=TRUE}



```


## SMART Question 4  
**Question4**

```{r Q4, message = FALSE, warning = FALSE, include=TRUE}



```



# Results 

## SMART Question1: 
- 


## SMART Question2:
- 


## SMART Question3: 
- 


## SMART Question4: 
- 


# Conclusion & Discussion



# Project Workflow Summary (structured outline)

1. Data acquisition & initial inspection
2. Data cleaning and variable derivation (document every transformation)
3. Exploratory Data Analysis (visual + summary statistics)
4. Hypothesis specification (SMART questions)
5. Statistical testing / modeling (with assumption checks)
6. Interpretation and robustness checks
7. Reporting: visualizations, tables for each SMART question, and final HTML


# Reference

Dataset Link:
https://health.data.ny.gov/Health/Hospital-Inpatient-Discharges-SPARCS-De-Identified/sf4k-39ay/about_data



<!-- End -->
