---
title: "Final Project: Examining the Relationship Between Demographics, Admission Types, and Clinical Outcomes in New York Inpatients (Niagara County)"
author: "Team 2: Simintaj Salehpour,Madeline Bumpus, Nazish Atta, Divya Mistry"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: flatly          # modern, clean blue theme
    highlight: tango       # gentle code highlight colors
    toc: true              # show table of contents
    toc_float: true        # floating TOC on the right
    number_sections: true  # adds section numbers
    df_print: paged        # pretty tables
    code_folding: hide     # hides code by default
    smooth_scroll: true    # smooth navigation
    self_contained: true   # all assets embedded
---
<style>
body {
  background-color: #f4f8fb; /* light blue-gray */
  color: #222;
  font-family: "Lato", "Helvetica Neue", sans-serif;
}
h1, h2, h3 {
  color: #004d80;
}
</style>
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(knitr)
library(kableExtra)
library(skimr)
library(naniar)
library(dplyr)
library(ezids)
theme_set(theme_minimal())
```

# Introduction

Our team chose to examine differences in **hospital length of stay (LOS)** across **insurance types** because it highlights an important intersection between healthcare access, cost, and patient outcomes. The relationship between insurance coverage and hospital utilization is central to understanding disparities in care delivery and efficiency within the U.S. healthcare system. 
We were particularly interested in comparing **Medicare**, which primarily serves older adults, and **Private Insurance**, which often provides broader access to care, to see how coverage type influences hospital resource use. Members of our team who have personally dealt with **Medicaid** and have seen firsthand how insurance status can impact the quality, timing, and length of medical care also find personal significance in this topic. We sought to better understand how insurance differences translate into quantifiable healthcare outcomes by combining our personal viewpoints with data-driven evidence through the analysis of actual hospital discharge data.


# Data Description

## Datasets used

This analysis uses data from the **New York State Department of Health (NYSDOH) Statewide Planning and Research Cooperative System (SPARCS)**, specifically the *Hospital Inpatient Discharges (SPARCS De-Identified)* dataset for **Niagara County, 2024**.
The dataset includes detailed records of all hospital discharges statewide, with variables describing patient demographics, diagnoses, procedures, hospital facility characteristics, insurance/payment type, admission type, and clinical outcomes. Each row represents one patient discharge, allowing for analysis of trends in hospital length of stay, cost, and patient mix.
The full dataset contains over **2 million records** across New York State; for this project, a subset of approximately **6,913 discharges from Niagara County** was used to maintain manageability and statistical power. This dataset is publicly available through the **NY Open Data Portal** and is widely used in healthcare policy, public health, and data science research.




```{r Load Dataset, include=T, results='markup'}
# Load CSV file
data <- read.csv("Hospital_Inpatient_Discharges_2024.csv")

#Clean column names to avoid backticks

names(data) <- make.names(names(data))

cat("Dataset dimensions: Rows:", nrow(data), "Columns:", ncol(data), "\n")
#glimpse(data)

```


## Missing data
- Report missingness by variable:
We check if there is a missing value or not. For each column check the missing value.

```{r Missing Value, include=T, results = "markup"}

# --- 1. Summary of missing values ---
missing_summary <- data %>% summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "column(variable)", values_to = "n_missing") %>% arrange(desc(n_missing))

#missing_summary

# --- 2. Visualize missingness per column ---
gg_miss_var(data) + labs(title = "Variables VS Number of Missing Values") +
  theme( plot.title = element_text(face = "bold", color = "darkblue"))

# --- 3. Check if there are any missing values ---
clean_data <- data
if(any(is.na(data))) {
  
  cat("Missing values detected!\n")
  total_missing <- sum(is.na(data))
  total_cells <- nrow(data) * ncol(data)
  
  # Remove rows if missing values are few (<5% of total data)
  if(total_missing / total_cells < 0.05) {
    clean_data <- na.omit(data)
    cat("Rows with missing values removed.\n")
  } 
  else {
    # Impute numeric columns with median
    for(col in names(clean_data)) {
      if(is.numeric(clean_data[[col]])) {          # Check if the column is numeric
          clean_data[[col]][is.na(clean_data[[col]])] <- median(clean_data[[col]], na.rm = TRUE)  
      }
    }
    cat("Missing values in numeric columns replaced with median.\n")
  }
  
} else {
  cat("No missing values detected. Dataset is clean.\n")
}


```


## Remove Duplicates
- Report Duplicates by row:
 Check to remove duplicates:
```{r Remove Duplicates, include=T, results = "markup"}
dup_count <- sum(duplicated(clean_data))
cat("Number of duplicate rows:", dup_count, "\n")

# Remove duplicates
clean_data <- clean_data %>% distinct()
```
# SMART Questions

The team developed the following **SMART research questions** based on data availability, clinical reasoning, and literature review. These focus on measurable outcomes—length of stay and costs—to analyze healthcare disparities.


1. What factors significantly increase the likelihood of a longer hospital stay (>3 days)?

2. What combination of patient demographics, admission type, and comorbidities predicts the
lowest hospital charges for Medicaid vs. Private Insurance patients in 2024?

3. Which combination of patient age group, severity of illness, and
payer type predicts the lowest hospital readmission risk?

4. Which factors predict in-hospital mortality, and how accurately can we classify high-risk patients?

## SMART Question 1  
**What factors significantly increase the likelihood of a long hospital stay (>3 days)?**



```{r Q1, message = FALSE, warning = FALSE, include=TRUE}
# What factors increase long LOS (>3 days)?

library(dplyr)
library(caret)
library(broom)
library(ggplot2)
library(randomForest)

# 1. Use the cleaned data 

df <- clean_data %>%
  # Create binary outcome
  mutate(
    LongStay = ifelse(Length.of.Stay > 3, 1, 0),

    # Insurance grouping
    Insurance_Group = case_when(
      Payment.Typology.1 == "Medicare" |
        Payment.Typology.2 == "Medicare" |
        Payment.Typology.3 == "Medicare" ~ "Medicare",

      Payment.Typology.1 == "Medicaid" |
        Payment.Typology.2 == "Medicaid" |
        Payment.Typology.3 == "Medicaid" ~ "Medicaid",

      Payment.Typology.1 %in% c("Blue Cross/Blue Shield", "Private Health Insurance") |
        Payment.Typology.2 %in% c("Blue Cross/Blue Shield", "Private Health Insurance") |
        Payment.Typology.3 %in% c("Blue Cross/Blue Shield", "Private Health Insurance") ~ "Private Insurance",

      TRUE ~ "Other"
    ),

    # Convert variables to factors
    LongStay = factor(LongStay, levels = c(0,1), labels = c("Short","Long")),
    Insurance_Group = factor(Insurance_Group),
    Age.Group = factor(Age.Group),
    Type.of.Admission = factor(Type.of.Admission),
    APR.Severity.of.Illness.Code = factor(APR.Severity.of.Illness.Code),
    Gender = factor(Gender)
  ) %>%
  select(LongStay, Insurance_Group, Age.Group, Type.of.Admission,
         APR.Severity.of.Illness.Code, Gender) %>%
  na.omit()

cat("Final modeling dataset rows:", nrow(df), "\n")

# Remove rows with 'Not Available' admission type to avoid new levels in test set
df <- df %>% filter(Type.of.Admission != "Not Available")

# Make sure Length.of.Stay is numeric
clean_data <- clean_data %>%
  mutate(Length.of.Stay = as.numeric(Length.of.Stay))
# Check distribution of Length of Stay
ggplot(clean_data, aes(x = Length.of.Stay)) +
  geom_histogram(binwidth = 1, fill = "#4a7ebb", color = "white") +
  labs(title = "Distribution of Length of Stay",
       x = "Length of Stay (Days)",
       y = "Count")

# Boxplot of LOS by Insurance Type (raw, pre-filter)
ggplot(clean_data, aes(x = Payment.Typology.1, y = Length.of.Stay)) +
  geom_boxplot(fill = "#87bdd8") +
  coord_flip() +
  labs(title = "Length of Stay by Insurance Type – Primary Payer",
       x = "Insurance Type",
       y = "Length of Stay")

# Check counts by insurance grouping
clean_data %>%
  count(Payment.Typology.1) %>%
  arrange(desc(n)) %>%
  kable(caption = "Counts of Insurance Types (Primary Payer)") %>%
  kable_styling(full_width = FALSE)

# LOS by Patient Age Group
ggplot(clean_data, aes(x = Age.Group, y = Length.of.Stay)) +
  geom_boxplot(fill = "#b3d9a8") +
  labs(title = "Length of Stay by Age Group",
       x = "Age Group",
       y = "Length of Stay (Days)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# LOS by Type of Admission
ggplot(clean_data, aes(x = Type.of.Admission, y = Length.of.Stay)) +
  geom_boxplot(fill = "#f3c677") +
  labs(title = "Length of Stay by Admission Type",
       x = "Admission Type",
       y = "Length of Stay (Days)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Crosstab: proportion of long stays by admission type
clean_data %>%
  mutate(LongStay = ifelse(Length.of.Stay > 3, "Long", "Short")) %>%
  count(Type.of.Admission, LongStay) %>%
  group_by(Type.of.Admission) %>%
  mutate(prop = n/sum(n)) %>%
  kable(caption = "Proportion of Long vs Short Stays by Admission Type") %>%
  kable_styling(full_width = FALSE)

# Checking severity distribution
ggplot(clean_data, aes(x = APR.Severity.of.Illness.Code)) +
  geom_bar(fill = "#d77fa1") +
  labs(title = "Severity of Illness Distribution",
       x = "APR Severity Code",
       y = "Number of Patients")
# 2. Train/Test Split
set.seed(123)
index <- createDataPartition(df$LongStay, p = 0.8, list = FALSE)
train <- df[index, ]
test <- df[-index, ]

# 3. Logistic Regression Model
log_model <- glm(
  LongStay ~ Insurance_Group + Age.Group + Type.of.Admission +
    APR.Severity.of.Illness.Code + Gender,
  data = train,
  family = binomial
)

summary(log_model)

# 4. Odds Ratios
odds <- tidy(log_model, exponentiate = TRUE, conf.int = TRUE)
kable(odds, caption = "Odds Ratios for Predictors of Long Hospital Stay")

# 5. Model Evaluation
pred_prob <- predict(log_model, newdata = test, type = "response")
pred_class <- ifelse(pred_prob > 0.5, "Long", "Short") |> factor(levels = c("Short","Long"))

conf_mat <- confusionMatrix(pred_class, test$LongStay, positive = "Long")

print(conf_mat)

cat("Model Accuracy:", round(conf_mat$overall["Accuracy"], 3), "\n")

# 6. Visualization of Significant Predictors
odds %>%
  filter(p.value < 0.05) %>%
  ggplot(aes(x = reorder(term, estimate), y = estimate)) +
  geom_col(fill = "#2c7fb8") +
  coord_flip() +
  labs(
    title = "Significant Predictors of Long Hospital Stay (>3 Days)",
    x = "Predictor",
    y = "Odds Ratio (Exp(Beta))"
  ) +
  theme_minimal()


# 7. Random Forest model to predict LongStay

cat("\n--- Random Forest model ---\n")

set.seed(123)
rf_fit <- randomForest(
  LongStay ~ .,
  data = train,
  ntree = 500,
  importance = TRUE
)

print(rf_fit)

rf_pred <- predict(rf_fit, newdata = test)
rf_cm <- confusionMatrix(rf_pred, test$LongStay, positive = "Long")
print(rf_cm)
cat("Random Forest Accuracy:", round(rf_cm$overall["Accuracy"], 3), "\n")

# Variable importance plot
varImpPlot(rf_fit, main = "Variable Importance for Predicting Long Hospital Stay")


# 8. ROC Curves & AUC Comparison

library(pROC)

cat("\n--- ROC Curve and AUC Comparison ---\n")

# --- Logistic Regression ROC ---
logit_prob <- predict(log_model, newdata = test, type = "response")

roc_logit <- roc(
  response = test$LongStay,
  predictor = logit_prob,
  levels = c("Short", "Long"),
  direction = "<"
)

logit_auc <- auc(roc_logit)
cat("Logistic Regression AUC:", round(logit_auc, 3), "\n")

# --- Random Forest ROC ---
rf_prob <- predict(rf_fit, newdata = test, type = "prob")[, "Long"]

roc_rf <- roc(
  response = test$LongStay,
  predictor = rf_prob,
  levels = c("Short", "Long"),
  direction = "<"
)

rf_auc <- auc(roc_rf)
cat("Random Forest AUC:", round(rf_auc, 3), "\n")

# --- Plot both ROC curves ---
plot(
  roc_logit, col = "blue", lwd = 2,
  main = "ROC Comparison: Logistic Regression vs Random Forest"
)
plot(
  roc_rf, col = "red", lwd = 2, add = TRUE
)

legend(
  "bottomright",
  legend = c(
    paste("Logistic Regression (AUC =", round(logit_auc, 3), ")"),
    paste("Random Forest (AUC =", round(rf_auc, 3), ")")
  ),
  col = c("blue", "red"),
  lwd = 2
)

```


## SMART Question 2  
**What combination of patient demographics, admission type, and comorbidities predicts the
lowest hospital charges for Medicaid vs. Private Insurance patients in 2024?**

```{r Q2, message = FALSE, warning = FALSE, include=TRUE}

library(dplyr)
library(forcats)
library(ggplot2)
library(randomForest)

# Data cleaning
clean_data <- clean_data %>%
  mutate(
    Total.Charges = as.numeric(gsub("[^0-9\\.]", "", Total.Charges)),
    Length.of.Stay = as.numeric(as.character(Length.of.Stay))
  ) %>%
  filter(!is.na(Total.Charges), is.finite(Total.Charges), Total.Charges > 0)



# Filter Insurance types
medicaid_pattern <- "Medicaid|federal/state/local/va"
private_pattern  <- "Blue Cross/Blue Shield|Private Health Insurance|Miscellaneous/Other"

hospital_filtered <- clean_data %>%
  mutate(
    Insurance = case_when(
      if_any(starts_with("Payment.Typology"), ~ str_detect(., medicaid_pattern)) ~ "Medicaid",
      if_any(starts_with("Payment.Typology"), ~ str_detect(., private_pattern))  ~ "Private",
      TRUE ~ "Other"
    )
  ) %>%
  filter(Insurance %in% c("Medicaid", "Private"))


# Keep relevant columns
columns_to_keep <- c(
  "Age.Group", "Gender","Race", "Length.of.Stay","Type.of.Admission",
  "Emergency.Department.Indicator","APR.Medical.Surgical.Description",
  "APR.Severity.of.Illness.Description","APR.DRG.Code","CCSR.Diagnosis.Description",
  "Insurance","Total.Charges"
)
hospital_final <- hospital_filtered[, columns_to_keep]

# Convert low frequency levels to "Other" and handle NA
factors_to_fix <- c(
  "Age.Group","Gender","Race","Type.of.Admission",
  "Emergency.Department.Indicator","APR.Severity.of.Illness.Description","CCSR.Diagnosis.Description")

for(f in factors_to_fix){
  hospital_final[[f]] <- fct_lump_min(as.factor(hospital_final[[f]]), min = 20, other_level = "Other")
  hospital_final[[f]] <- fct_explicit_na(hospital_final[[f]], na_level = "Other")
}
cat("After labling Insurances as Medicaid and Private and remove other kinds like Medicare and self-payment \nStructure of final filtered dataset with demographics, admission type, and comorbidities variables to predicts the lowest hospital charges for Medicaid vs. Private Insurance: ")
glimpse(hospital_final)

# Remove rows with any NA in predictors
predictors <- c(
  "Age.Group","Gender","Race","Type.of.Admission","Length.of.Stay",
  "Emergency.Department.Indicator","APR.Severity.of.Illness.Description","CCSR.Diagnosis.Description")
hospital_final <- hospital_final %>% filter(!if_any(all_of(predictors), is.na))


# Train/test split function with safe factor alignment
split_train_test <- function(df, train_ratio = 0.8) {
  set.seed(123)
  n <- nrow(df)
  train_idx <- sample(1:n, size = floor(train_ratio * n))
  train <- df[train_idx, ]
  test  <- df[-train_idx, ]
  
  for(f in names(train)[sapply(train, is.factor)]) {
    test[[f]] <- factor(test[[f]], levels = levels(train[[f]]))
    test[[f]] <- fct_explicit_na(test[[f]], na_level = "Other")
  }
  
  list(train = train, test = test)
}

# Split by Insurance
Medicaid_df <- hospital_final %>% filter(Insurance == "Medicaid")
Private_df  <- hospital_final %>% filter(Insurance == "Private")

medicaid_split <- split_train_test(Medicaid_df)
private_split  <- split_train_test(Private_df)

medicaid_split$train$logCharges <- log(medicaid_split$train$Total.Charges + 1)
medicaid_split$test$logCharges  <- log(medicaid_split$test$Total.Charges + 1)

private_split$train$logCharges <- log(private_split$train$Total.Charges + 1)
private_split$test$logCharges  <- log(private_split$test$Total.Charges + 1)
# Random Forest models

set.seed(123)
rf_medicaid <- randomForest(
  logCharges  ~ Age.Group + Gender + Race +
    Type.of.Admission + Length.of.Stay +
    Emergency.Department.Indicator +
    APR.Severity.of.Illness.Description +
    CCSR.Diagnosis.Description,
  data = medicaid_split$train,
  ntree = 500,
  mtry = 3,
  importance = TRUE
)
rf_medicaid
# R² calculation
pred_log <- predict(rf_medicaid, medicaid_split$test)
real_log <- medicaid_split$test$logCharges
# R2 on log scale
# R2_log <- 1 - sum((pred_log - real_log)^2) / sum((real_log - mean(real_log))^2)
# cat("Medicaid:R2 (log scale):", R2_log, "\n")

# R2 on real scale
pred_real <- exp(pred_log) - 1
real_real <- medicaid_split$test$Total.Charges

R2_real <- 1 - sum((pred_real - real_real)^2) / sum((real_real - mean(real_real))^2)
cat("Medicaid: R2 (real scale):", R2_real, "\n")



rf_private <- randomForest(
  logCharges ~ Age.Group + Gender + Race +
    Type.of.Admission + Length.of.Stay +
    Emergency.Department.Indicator +
    APR.Severity.of.Illness.Description +
    CCSR.Diagnosis.Description,
  data = private_split$train,
  ntree = 500,
  mtry = 3,
  importance = TRUE
)
rf_private
# R² calculation
pred_log <- predict(rf_private, private_split$test)
real_log <- private_split$test$logCharges
# R2 on log scale
# R2_log <- 1 - sum((pred_log - real_log)^2) / sum((real_log - mean(real_log))^2)
# cat("Private:R2 (log scale):", R2_log, "\n")

# R2 on real scale
pred_real <- exp(pred_log) - 1
real_real <- private_split$test$Total.Charges

R2_real <- 1 - sum((pred_real - real_real)^2) / sum((real_real - mean(real_real))^2)
cat("Private: R2 (real scale):", R2_real, "\n")



top_features <- function(rf_model, n = 5){
  imp <- importance(rf_model)
  imp_df <- data.frame(
    Feature = rownames(imp),
    Importance = imp[, "IncNodePurity"]
  ) %>%
    arrange(desc(Importance))
  head(imp_df$Feature, n)
}

# Feature importance plots
plot_feature_importance <- function(rf_model, title = "Feature Importance") {
  imp <- importance(rf_model)
  imp_df <- data.frame(
    Feature = rownames(imp),
    Importance = imp[, "IncNodePurity"]
  ) %>% arrange(desc(Importance))
  
  ggplot(imp_df, aes(x = reorder(Feature, Importance), y = Importance)) +
    geom_col(fill = "steelblue") +
    coord_flip() +
    theme_minimal() +
    labs(title = title, x = "Feature", y = "Importance (IncNodePurity)")
}

# Medicaid
plot_feature_importance(rf_medicaid, title = "Feature Importance (Medicaid)")

# Private
plot_feature_importance(rf_private, title = "Feature Importance (Private Insurance)")


# Predictions
medicaid_split$test$pred_rf <- predict(rf_medicaid, newdata = medicaid_split$test)
private_split$test$pred_rf  <- predict(rf_private, newdata = private_split$test)

# Top 10 lowest predicted charges
lowest_medicaid <- medicaid_split$test %>% arrange(pred_rf) %>% head(10)
lowest_private  <- private_split$test %>% arrange(pred_rf) %>% head(10)

cat("Top 10 lowest predicted charges (Medicaid):\n")
print(lowest_medicaid)
cat("The lowest predicted charges for Medicaid patients occur for youngers and newborns, with 1-day hospital stays, minor illness severity, non-emergency admissions, and conditions such as live births for newborns and OPIOID-RELATED DISORDERS (30-49) years old.
Admission types include Elective, Urgent, and Newborn, but Newborn cases dominate the lowest charge group.")


cat("Top 10 lowest predicted charges (Private):\n")
print(lowest_private)
cat("The lowest predicted hospital charges for Private Insurance patients occur for newborn babies (LIVEBORN) with 1-day or 2-days stays, minor illness severity, and non-ED admission. All 10 cases have Type of Admission = Newborn, which is the strongest single predictor of low charges in the Private group.")



library(caret)

# Function to create LowCharge, fit logistic regression, and evaluate

evaluate_lowcharge <- function(df_split, median_value, group_name) {
  
  # Create binary outcome
  df_split$train$LowCharge <- ifelse(df_split$train$Total.Charges <= median_value, 1, 0)
  df_split$test$LowCharge  <- ifelse(df_split$test$Total.Charges <= median_value, 1, 0)
  
  # Fit logistic regression
  logit_model <- glm(LowCharge ~ Age.Group + Gender + Race + Type.of.Admission +
                       Length.of.Stay + Emergency.Department.Indicator +
                       APR.Severity.of.Illness.Description +
                       CCSR.Diagnosis.Description,
                     data = df_split$train, family = binomial)
  
  # Predict on test set
  pred_class <- ifelse(predict(logit_model, newdata = df_split$test, type = "response") > 0.5, 1, 0)
  
  # Confusion matrix
  cm <- confusionMatrix(factor(pred_class), factor(df_split$test$LowCharge), positive="1")
  
  cat("\n---", group_name, "---\n")
  print(cm)
  
  return(list(model = logit_model, confusion = cm))
}

median_medicaid <- median(Medicaid_df$Total.Charges)
median_private  <- median(Private_df$Total.Charges)

result_medicaid <- evaluate_lowcharge(medicaid_split, median_medicaid, "Medicaid")
result_private  <- evaluate_lowcharge(private_split, median_private, "Private")



```



## SMART Question 3  
**Which combination of patient age group, severity of illness, and
payer type predicts the lowest hospital readmission risk?**

```{r Q3, message = FALSE, warning = FALSE, include=TRUE}
# ----------------------------
# Load libraries
# ----------------------------
library(dplyr)
library(ggplot2)
library(caret)
library(pROC)
library(reshape2)
library(tidyr)
library(kableExtra)
library(scales)

# ----------------------------
# 1. Load and clean dataset
# ----------------------------
data <- read.csv("Hospital_Inpatient_Discharges_2024.csv")
names(data) <- make.names(names(data))  # clean column names
cat("Dataset dimensions: Rows:", nrow(data), "Columns:", ncol(data), "\n")

# ----------------------------
# 2. Check missing values
# ----------------------------
missing_summary <- data %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "n_missing") %>%
  arrange(desc(n_missing))
missing_summary

# ----------------------------
# 3. Keep useful variables & create proxy readmission
# ----------------------------
readmit_data <- data %>%
  select(Age.Group, Gender, Zip.Code, APR.Severity.of.Illness.Description, Payment.Typology.1) %>%
  filter(
    !is.na(Age.Group),
    !is.na(Gender),
    !is.na(Zip.Code),
    !is.na(APR.Severity.of.Illness.Description),
    !is.na(Payment.Typology.1)
  )

# Proxy readmission: multiple records per patient (Zip + Age + Gender)
readmit_data <- readmit_data %>%
  group_by(Zip.Code, Age.Group, Gender) %>%
  mutate(Readmission = ifelse(n() > 1, 1, 0)) %>%
  ungroup()

readmit_data$Readmission <- factor(readmit_data$Readmission, levels = c(0,1), labels = c("No", "Yes"))

# ----------------------------
# Check class balance
# ----------------------------
cat("Class distribution:\n")
print(table(readmit_data$Readmission))

# ----------------------------
# 4. Stratified Train/Test split
# ----------------------------
set.seed(123)
index <- createDataPartition(readmit_data$Readmission, p = 0.7, list = FALSE)
train <- readmit_data[index, ]
test <- readmit_data[-index, ]

cat("Train class distribution:\n")
print(table(train$Readmission))
cat("Test class distribution:\n")
print(table(test$Readmission))

# ----------------------------
# 5. Logistic Regression Model
# ----------------------------
logit_model <- glm(
  Readmission ~ Age.Group + APR.Severity.of.Illness.Description + Payment.Typology.1,
  data = train,
  family = binomial
)
summary(logit_model)

# ----------------------------
# 6. Predictions + ROC Curve (safe)
# ----------------------------
pred_prob <- predict(logit_model, newdata = test, type = "response")

# Only compute ROC if test has both classes
if(length(unique(test$Readmission)) == 2) {
  roc_obj <- roc(test$Readmission, pred_prob)
  auc_value <- auc(roc_obj)
  plot(
    roc_obj,
    col = "#0066CC",
    lwd = 2,
    main = paste("ROC Curve for Readmission Risk (AUC =", round(auc_value, 3), ")")
  )
} else {
  cat("ROC curve cannot be computed: test set has only one class.\n")
}

# ----------------------------
# 7. Confusion Matrix
# ----------------------------
cutoff <- 0.10
pred_class <- ifelse(pred_prob >= cutoff, "Yes", "No")
pred_class <- factor(pred_class, levels = c("No", "Yes"))

conf_mat <- confusionMatrix(pred_class, test$Readmission, positive = "Yes")
conf_mat

# ----------------------------
# 8. Visual 1: Readmission Rate by Age & Severity
# ----------------------------
ggplot(readmit_data, aes(x = Age.Group, fill = APR.Severity.of.Illness.Description)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = percent_format()) +
  labs(
    title = "Readmission Proportions by Age Group and Severity of Illness",
    x = "Age Group",
    y = "Proportion Readmitted",
    fill = "Severity Level"
  ) +
  theme_minimal(base_size = 14) +
  scale_fill_brewer(palette = "Blues")

# ----------------------------
# 9. Visual 2: Readmission Rate by Payer
# ----------------------------
readmission_by_payer <- readmit_data %>%
  group_by(Payment.Typology.1) %>%
  summarise(Readmission_Rate = mean(Readmission == "Yes")) %>%
  arrange(Readmission_Rate)

ggplot(readmission_by_payer, aes(x = reorder(Payment.Typology.1, Readmission_Rate), y = Readmission_Rate)) +
  geom_col(fill = "#0066CC") +
  geom_text(aes(label = percent(Readmission_Rate, accuracy = 0.1)), vjust = -0.3, size = 4) +
  labs(
    title = "Readmission Risk by Payer Type",
    x = "Payer Type",
    y = "Readmission Rate"
  ) +
  theme_minimal(base_size = 14)

# ----------------------------
# 10. Visual 3: Heatmap by Age, Severity, and Payer
# ----------------------------
readmission_heatmap <- readmit_data %>%
  group_by(Age.Group, APR.Severity.of.Illness.Description, Payment.Typology.1) %>%
  summarise(Readmission_Rate = mean(Readmission == "Yes")) %>%
  ungroup()

heatmap_data <- dcast(
  readmission_heatmap,
  Age.Group + APR.Severity.of.Illness.Description ~ Payment.Typology.1,
  value.var = "Readmission_Rate",
  fill = 0
)

heatmap_long <- melt(heatmap_data, id.vars = c("Age.Group", "APR.Severity.of.Illness.Description"))

ggplot(heatmap_long, aes(x = variable, y = interaction(Age.Group, APR.Severity.of.Illness.Description), fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "white", high = "#0066CC", name = "Readmission Rate") +
  labs(
    title = "Heatmap of Readmission Risk by Age, Severity, and Payer",
    x = "Payer Type",
    y = "Age Group + Severity"
  ) +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# ----------------------------
# 11. Summary Table
# ----------------------------
readmission_summary <- readmit_data %>%
  group_by(Age.Group, APR.Severity.of.Illness.Description, Payment.Typology.1) %>%
  summarise(Readmission_Rate = mean(Readmission == "Yes")) %>%
  arrange(Readmission_Rate)

readmission_summary %>%
  kable() %>%
  kable_styling(full_width = FALSE)


```


## SMART Question 4  
**Which factors predict in-hospital mortality, and how accurately can we classify high-risk patients?
**

```{r Q4, message = FALSE, warning = FALSE, include=TRUE}

# --------------------------------------------------
# SMART Question 4:
# Which factors predict in-hospital mortality, and
# how accurately can we classify high-risk patients?

#Goal: The goal was to find out which factors increase a patient’s chance of dying in the hospital and how accurately those high-risk patients can be identified.
# -------------------------------------------------
library(dplyr)
library(ggplot2)
library(caret)
library(pROC)


set.seed(123)

# 1. Create binary outcome: 1 = died in hospital, 0 = survived
clean_data$Mortality <- ifelse(clean_data$Patient.Disposition == "Expired", 1L, 0L)
#2 missing values
q4_data <- clean_data %>%
  filter(
    !is.na(APR.Risk.of.Mortality),
    !is.na(APR.Severity.of.Illness.Description),
    APR.Risk.of.Mortality != "Undetermined",
    APR.Severity.of.Illness.Description != "Undetermined",
    !is.na(Length.of.Stay)
  ) %>%
  select(
    Mortality,
    Age.Group,
    Type.of.Admission,
    Emergency.Department.Indicator,
    APR.Severity.of.Illness.Description,
    APR.Risk.of.Mortality,
    Length.of.Stay
  )

# 3. Set variable types 

q4_data$Mortality      <- as.integer(q4_data$Mortality)
q4_data$MortalityClass <- factor(q4_data$Mortality,
                                 levels = c(0, 1),
                                 labels = c("Survived", "Died"))

q4_data$Age.Group                      <- droplevels(as.factor(q4_data$Age.Group))
q4_data$Type.of.Admission              <- droplevels(as.factor(q4_data$Type.of.Admission))
q4_data$Emergency.Department.Indicator <- droplevels(as.factor(q4_data$Emergency.Department.Indicator))
q4_data$APR.Severity.of.Illness.Description <- droplevels(as.factor(q4_data$APR.Severity.of.Illness.Description))
q4_data$APR.Risk.of.Mortality          <- droplevels(as.factor(q4_data$APR.Risk.of.Mortality))
q4_data$Length.of.Stay                 <- as.numeric(q4_data$Length.of.Stay)


# 4. Train/Test split

index_q4 <- createDataPartition(q4_data$MortalityClass, p = 0.7, list = FALSE)
train_q4 <- q4_data[index_q4, ]
test_q4  <- q4_data[-index_q4, ]

# 5. Logistic regression model

logit_q4 <- glm(
  Mortality ~ Age.Group +
    Type.of.Admission +
    Emergency.Department.Indicator +
    APR.Severity.of.Illness.Description +
    APR.Risk.of.Mortality +
    Length.of.Stay,
  data   = train_q4,
  family = binomial
)

summary(logit_q4)


# 6. Predicted probabilities on TEST set 

pred_prob <- predict(logit_q4, newdata = test_q4, type = "response")

# 7. ROC curve and AUC 

roc_q4 <- roc(test_q4$Mortality, pred_prob)
auc_q4 <- auc(roc_q4)

plot(roc_q4,
     col  = "#0066CC",
     lwd  = 2,
     main = paste("ROC Curve for In-Hospital Mortality (AUC =",
                  round(auc_q4, 3), ")"))


# 8. Classify "high-risk" patients at cutoff = 0.10

cutoff <- 0.10
pred_class_num <- ifelse(pred_prob >= cutoff, 1L, 0L)
pred_class_fac <- factor(pred_class_num,
                         levels = c(0, 1),
                         labels = c("Survived", "Died"))

# Confusion matrix on test set
conf_mat_q4 <- confusionMatrix(
  pred_class_fac,
  test_q4$MortalityClass,
  positive = "Died"
)

conf_mat_q4

# Extract key metrics
accuracy    <- conf_mat_q4$overall["Accuracy"]
sensitivity <- conf_mat_q4$byClass["Sensitivity"]
specificity <- conf_mat_q4$byClass["Specificity"]
ppv         <- conf_mat_q4$byClass["Pos Pred Value"]
npv         <- conf_mat_q4$byClass["Neg Pred Value"]

# 9. Visualization: Mortality by APR Risk 

ggplot(train_q4,
       aes(x = APR.Risk.of.Mortality,
           fill = MortalityClass)) +
  geom_bar(position = "fill", color = "black") +
  scale_fill_manual(values = c("Survived" = "#FF9933",
                               "Died"      = "#0066CC")) +
  labs(
    title = "Proportion of In-Hospital Deaths by APR Risk of Mortality",
    x     = "APR Risk of Mortality",
    y     = "Proportion within Risk Group",
    fill  = "Outcome"
  ) +
  theme_bw() +
  theme(
    plot.title   = element_text(hjust = 0.5, face = "bold", size = 14),
    axis.title   = element_text(size = 12),
    axis.text    = element_text(size = 10),
    legend.title = element_text(size = 11),
    legend.text  = element_text(size = 10),
    legend.position = "right"
  )


# 10. Calculate mortality rate for each category of each predictor

mortality_summary <- bind_rows(
  q4_data %>%
    group_by(APR.Risk.of.Mortality) %>%
    summarise(mortality_rate = mean(Mortality)) %>%
    mutate(Factor = "APR Risk of Mortality",
           Category = APR.Risk.of.Mortality),

  q4_data %>%
    group_by(APR.Severity.of.Illness.Description) %>%
    summarise(mortality_rate = mean(Mortality)) %>%
    mutate(Factor = "APR Severity of Illness",
           Category = APR.Severity.of.Illness.Description),

  q4_data %>%
    group_by(Type.of.Admission) %>%
    summarise(mortality_rate = mean(Mortality)) %>%
    mutate(Factor = "Admission Type",
           Category = Type.of.Admission),

  q4_data %>%
    group_by(Age.Group) %>%
    summarise(mortality_rate = mean(Mortality)) %>%
    mutate(Factor = "Age Group",
           Category = Age.Group)
)
mortality_summary$Category <- as.factor(mortality_summary$Category)

# one small chart per factor
ggplot(mortality_summary,
       aes(x = Category, y = mortality_rate * 100, fill = Factor)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ Factor, scales = "free_x") +
  labs(
    title = "Comparison of Mortality Rates Across Predictors",
    x = NULL,
    y = "Mortality Rate (%)"
  ) +
  theme_bw() +
  theme(
    strip.text = element_text(face = "bold", size = 12),
    axis.text.x = element_text(angle = 30, hjust = 1)
  )


# 11. summary

cat("\n--- SMART Question 4 Summary ---\n")
cat("AUC:", round(auc_q4, 3), "\n")
cat("Accuracy:", round(accuracy, 3), "\n")
cat("Sensitivity:", round(sensitivity, 3), "\n")
cat("Specificity:", round(specificity, 3), "\n")
cat("PPV:", round(ppv, 3), "\n")
cat("NPV:", round(npv, 3), "\n")


```


# Analysis 

## SMART Question1:
Existing literature consistently shows that hospital length of stay is influenced by demographic variables (such as age), insurance status, admission type, and clinical severity—factors that are also strong predictors of hospital resource use and patient outcomes. These findings directly informed the development of our SMART Question 1, which focuses on identifying predictors of long hospital stays (>3 days). During the exploratory data analysis (EDA), we observed similar patterns in the SPARCS dataset: Medicare and Medicaid patients showed notably longer stays, emergency admissions exhibited wider LOS variability, and higher APR severity levels corresponded with longer recovery times. These early patterns helped refine our modeling approach by confirming which variables were most relevant and should be included in the logistic regression and random forest models.

We selected logistic regression because our outcome (whether a patient stayed longer than 3 days) was a binary classification problem. Logistic regression provides interpretable coefficients and odds ratios, which allow us to identify which factors significantly increase the likelihood of a long stay. We also used a random forest classifier as a complementary, non-linear model to check whether more complex interactions or variable importance patterns supported the regression findings. Logistic regression offered better interpretability, while random forest helped validate which predictors mattered most.

Using the model, we can predict a patient’s probability of staying more than 3 days based on characteristics such as:

- Admission type (e.g., emergency admissions predic longer stays)

- Severity of illness (higher APR severity levels predict longer LOS)

- Insurance type (patients with Private Insurance or “Other” tend to have lower odds of long stays compared to Medicaid)

- Age group (older patients show increased likelihood, though the effect size is smaller) 

Our analysis found that admission type is the strongest predictor of long hospital stays. Patients admitted with an Emergency/Urgent label has much higher odds of extended stays,
while insurance type played a smaller but still meaningful role. Age, gender, and severity of illness showed little predictive power in the logistic model. Random Forest highlighted similar variables but performed slightly worse overall. Logistic regression provided clearer, more interpretable results. Overall, admission-related factors, not demographics, best explain who is likely to stay in the hospital longer.

Reliability is moderate but not perfect: Logistic regression and random forest achieved ~61–62% accuracy, only slightly above baseline. Specificity is very high, meaning the models are excellent at predicting short stays. Sensitivity is low, meaning the models struggle to detect true long stays: likely because long stays are harder to predict and may depend on unobserved clinical factors. ROC curves showed AUC values in the moderate range, consistent with partial predictive ability.

Several enhancements could strengthen predictive power:

1. Include clinical detail not captured in the dataset

- Comorbidities 

- Procedure type or surgical complexity

- Vital signs or lab values

- Medication use and complications

2. Use more advanced modeling approaches

- Gradient boosting (XGBoost, LightGBM)

- Elastic net logistic regression (better variable selection)

- SMOTE or balanced resampling to improve detection of long stays

3. Incorporate interaction terms

- For example: Emergency admission × high severity or Age × severity

4. Expand categorical groups

- Collapsing categories (like insurance) may hide meaningful subpatterns.

5. Use multiple years of SPARCS data

- A larger sample can improve model stability and help detect rare long-stay patterns.



## SMART Question2: ##
Based on the feature importance from the Random Forest models and Across both Medicaid and Private Insurance, Length of Stay, CCSR Diagnosis, Type of Admission, and APR Severity of Illness emerge as the most influential predictors of low hospital charges, with the lowest costs consistently linked to Newborn admissions, very short (1–2 day) stays, minor illness severity, and low-complexity diagnoses.

- 


## SMART Question3: ##
The analysis shows that readmission risk
is mainly driven by **severity of illness**, **age**, and **payer type**. Patients with higher severity levels and older age groups have notably higher readmission rates. Medicare and Medicaid patients also show increased risk compared to commercial payers. The model and visualizations consistently support these patterns. Overall, focusing follow-up care and support on high-severity, older, and publicly insured patients may help reduce avoidable readmissions.
- 


## SMART Question4: 
In-hospital mortality
was rare (~2–3%), so accuracy alone is not very informative. The model showed high specificity and NPV, meaning it reliably identified patients who survived while still detecting a meaningful portion of true deaths. APR Risk of Mortality and APR Severity of Illness were by far the strongest predictors—mortality rose from nearly 0% in Minor/Moderate risk groups to ~5% in Major risk and ~25% in Extreme risk categories. Older age (70+), and emergency or urgent admissions also increased mortality risk but to a lesser extent. Overall, clinical severity and APR risk scores drive mortality outcomes, and the logistic model provides a solid framework for distinguishing high-risk from low-risk patients.
- 



# Reference
New. (2025, October 8). Hospital Inpatient Discharges (SPARCS De-Identified): 2024. Ny.gov. https://health.data.ny.gov/Health/Hospital-Inpatient-Discharges-SPARCS-De-Identified/sf4k-39ay/about_data




<!-- End -->
