---
title: "Final Project: Examining the Relationship Between Demographics, Admission Types, and Clinical Outcomes in New York Inpatients (Niagara County)"
author: "Team 2: Simintaj Salehpour,Madeline Bumpus, Nazish Atta, Divya Mistry"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: flatly          # modern, clean blue theme
    highlight: tango       # gentle code highlight colors
    toc: true              # show table of contents
    toc_float: true        # floating TOC on the right
    number_sections: true  # adds section numbers
    df_print: paged        # pretty tables
    code_folding: hide     # hides code by default
    smooth_scroll: true    # smooth navigation
    self_contained: true   # all assets embedded
---
<style>
body {
  background-color: #f4f8fb; /* light blue-gray */
  color: #222;
  font-family: "Lato", "Helvetica Neue", sans-serif;
}
h1, h2, h3 {
  color: #004d80;
}
</style>
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(knitr)
library(kableExtra)
library(skimr)
library(naniar)
library(dplyr)
library(ezids)
theme_set(theme_minimal())
```

# Introduction

Our team chose to examine differences in **hospital length of stay (LOS)** across **insurance types** because it highlights an important intersection between healthcare access, cost, and patient outcomes. The relationship between insurance coverage and hospital utilization is central to understanding disparities in care delivery and efficiency within the U.S. healthcare system. 
We were particularly interested in comparing **Medicare**, which primarily serves older adults, and **Private Insurance**, which often provides broader access to care, to see how coverage type influences hospital resource use. Members of our team who have personally dealt with **Medicaid** and have seen firsthand how insurance status can impact the quality, timing, and length of medical care also find personal significance in this topic. We sought to better understand how insurance differences translate into quantifiable healthcare outcomes by combining our personal viewpoints with data-driven evidence through the analysis of actual hospital discharge data.


# Data Description

## Datasets used

This analysis uses data from the **New York State Department of Health (NYSDOH) Statewide Planning and Research Cooperative System (SPARCS)**, specifically the *Hospital Inpatient Discharges (SPARCS De-Identified)* dataset for **Niagara County, 2024**.
The dataset includes detailed records of all hospital discharges statewide, with variables describing patient demographics, diagnoses, procedures, hospital facility characteristics, insurance/payment type, admission type, and clinical outcomes. Each row represents one patient discharge, allowing for analysis of trends in hospital length of stay, cost, and patient mix.
The full dataset contains over **2 million records** across New York State; for this project, a subset of approximately **6,913 discharges from Niagara County** was used to maintain manageability and statistical power. This dataset is publicly available through the **NY Open Data Portal** and is widely used in healthcare policy, public health, and data science research.




```{r Load Dataset, include=T, results='markup'}
# Load CSV file
data <- read.csv("Hospital_Inpatient_Discharges_2024.csv")

#Clean column names to avoid backticks

names(data) <- make.names(names(data))

cat("Dataset dimensions: Rows:", nrow(data), "Columns:", ncol(data), "\n")
#glimpse(data)

```


## Missing data
- Report missingness by variable:
We check if there is a missing value or not. For each column check the missing value.

```{r Missing Value, include=T, results = "markup"}

# --- 1. Summary of missing values ---
missing_summary <- data %>% summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "column(variable)", values_to = "n_missing") %>% arrange(desc(n_missing))

#missing_summary

# --- 2. Visualize missingness per column ---
gg_miss_var(data) + labs(title = "Variables VS Number of Missing Values") +
  theme( plot.title = element_text(face = "bold", color = "darkblue"))

# --- 3. Check if there are any missing values ---
clean_data <- data
if(any(is.na(data))) {
  
  cat("Missing values detected!\n")
  total_missing <- sum(is.na(data))
  total_cells <- nrow(data) * ncol(data)
  
  # Remove rows if missing values are few (<5% of total data)
  if(total_missing / total_cells < 0.05) {
    clean_data <- na.omit(data)
    cat("Rows with missing values removed.\n")
  } 
  else {
    # Impute numeric columns with median
    for(col in names(clean_data)) {
      if(is.numeric(clean_data[[col]])) {          # Check if the column is numeric
          clean_data[[col]][is.na(clean_data[[col]])] <- median(clean_data[[col]], na.rm = TRUE)  
      }
    }
    cat("Missing values in numeric columns replaced with median.\n")
  }
  
} else {
  cat("No missing values detected. Dataset is clean.\n")
}


```


## Remove Duplicates
- Report Duplicates by row:
 Check to remove duplicates:
```{r Remove Duplicates, include=T, results = "markup"}
dup_count <- sum(duplicated(clean_data))
cat("Number of duplicate rows:", dup_count, "\n")

# Remove duplicates
clean_data <- clean_data %>% distinct()
```
# SMART Questions

The team developed the following **SMART research questions** based on data availability, clinical reasoning, and literature review. These focus on measurable outcomes—length of stay and costs—to analyze healthcare disparities.


1. What factors significantly increase the likelihood of a longer hospital stay (>3 days)?

2. **SMART Question2**

3. **SMART Question3**

4. **SMART Question4**



## SMART Question 1  
**What factors significantly increase the likelihood of a long hospital stay (>3 days)?**

```{r Q1, message = FALSE, warning = FALSE, include=TRUE}
# ---------------------------------------------------------
# Logistic Regression: What factors increase long LOS (>3 days)?
# Using cleaned dataset from your earlier steps
# ---------------------------------------------------------

library(dplyr)
library(caret)
library(broom)
library(ggplot2)

# 1. Use the cleaned data 

df <- clean_data %>%
  # Create binary outcome
  mutate(
    LongStay = ifelse(Length.of.Stay > 3, 1, 0),

    # Insurance grouping
    Insurance_Group = case_when(
      Payment.Typology.1 == "Medicare" |
        Payment.Typology.2 == "Medicare" |
        Payment.Typology.3 == "Medicare" ~ "Medicare",

      Payment.Typology.1 == "Medicaid" |
        Payment.Typology.2 == "Medicaid" |
        Payment.Typology.3 == "Medicaid" ~ "Medicaid",

      Payment.Typology.1 %in% c("Blue Cross/Blue Shield", "Private Health Insurance") |
        Payment.Typology.2 %in% c("Blue Cross/Blue Shield", "Private Health Insurance") |
        Payment.Typology.3 %in% c("Blue Cross/Blue Shield", "Private Health Insurance") ~ "Private Insurance",

      TRUE ~ "Other"
    ),

    # Convert variables to factors
    LongStay = factor(LongStay, levels = c(0,1), labels = c("Short","Long")),
    Insurance_Group = factor(Insurance_Group),
    Age.Group = factor(Age.Group),
    Type.of.Admission = factor(Type.of.Admission),
    APR.Severity.of.Illness.Code = factor(APR.Severity.of.Illness.Code),
    Gender = factor(Gender)
  ) %>%
  select(LongStay, Insurance_Group, Age.Group, Type.of.Admission,
         APR.Severity.of.Illness.Code, Gender) %>%
  na.omit()

cat("Final modeling dataset rows:", nrow(df), "\n")

df <- df %>% filter(Type.of.Admission != "Not Available")

# 2. Train/Test Split
set.seed(123)
index <- createDataPartition(df$LongStay, p = 0.8, list = FALSE)
train <- df[index, ]
test <- df[-index, ]

# 3. Logistic Regression Model
log_model <- glm(
  LongStay ~ Insurance_Group + Age.Group + Type.of.Admission +
    APR.Severity.of.Illness.Code + Gender,
  data = train,
  family = binomial
)

summary(log_model)

# 4. Odds Ratios
odds <- tidy(log_model, exponentiate = TRUE, conf.int = TRUE)
kable(odds, caption = "Odds Ratios for Predictors of Long Hospital Stay")

# 5. Model Evaluation
pred_prob <- predict(log_model, newdata = test, type = "response")
pred_class <- ifelse(pred_prob > 0.5, "Long", "Short") |> factor(levels = c("Short","Long"))

conf_mat <- confusionMatrix(pred_class, test$LongStay, positive = "Long")

print(conf_mat)

cat("Model Accuracy:", round(conf_mat$overall["Accuracy"], 3), "\n")

# 6. Visualization of Significant Predictors
odds %>%
  filter(p.value < 0.05) %>%
  ggplot(aes(x = reorder(term, estimate), y = estimate)) +
  geom_col(fill = "#2c7fb8") +
  coord_flip() +
  labs(
    title = "Significant Predictors of Long Hospital Stay (>3 Days)",
    x = "Predictor",
    y = "Odds Ratio (Exp(Beta))"
  ) +
  theme_minimal()




```


## SMART Question 2  
**Question1**

```{r Q2, message = FALSE, warning = FALSE, include=TRUE}



```



## SMART Question 3  
**Question3**

```{r Q3, message = FALSE, warning = FALSE, include=TRUE}



```


## SMART Question 4  
**Question4**

```{r Q4, message = FALSE, warning = FALSE, include=TRUE}

# --------------------------------------------------
# SMART Question 4:
# Which factors predict in-hospital mortality, and
# how accurately can we classify high-risk patients?
# -------------------------------------------------
library(dplyr)
library(ggplot2)
library(caret)
library(pROC)


set.seed(123)

# 1. Create binary outcome: 1 = died in hospital, 0 = survived
clean_data$Mortality <- ifelse(clean_data$Patient.Disposition == "Expired", 1L, 0L)
#2 missing values
q4_data <- clean_data %>%
  filter(
    !is.na(APR.Risk.of.Mortality),
    !is.na(APR.Severity.of.Illness.Description),
    APR.Risk.of.Mortality != "Undetermined",
    APR.Severity.of.Illness.Description != "Undetermined",
    !is.na(Length.of.Stay)
  ) %>%
  select(
    Mortality,
    Age.Group,
    Type.of.Admission,
    Emergency.Department.Indicator,
    APR.Severity.of.Illness.Description,
    APR.Risk.of.Mortality,
    Length.of.Stay
  )

# 3. Set variable types 

q4_data$Mortality      <- as.integer(q4_data$Mortality)
q4_data$MortalityClass <- factor(q4_data$Mortality,
                                 levels = c(0, 1),
                                 labels = c("Survived", "Died"))

q4_data$Age.Group                      <- droplevels(as.factor(q4_data$Age.Group))
q4_data$Type.of.Admission              <- droplevels(as.factor(q4_data$Type.of.Admission))
q4_data$Emergency.Department.Indicator <- droplevels(as.factor(q4_data$Emergency.Department.Indicator))
q4_data$APR.Severity.of.Illness.Description <- droplevels(as.factor(q4_data$APR.Severity.of.Illness.Description))
q4_data$APR.Risk.of.Mortality          <- droplevels(as.factor(q4_data$APR.Risk.of.Mortality))
q4_data$Length.of.Stay                 <- as.numeric(q4_data$Length.of.Stay)

# 4. Train/Test split

index_q4 <- createDataPartition(q4_data$MortalityClass, p = 0.7, list = FALSE)
train_q4 <- q4_data[index_q4, ]
test_q4  <- q4_data[-index_q4, ]

# 5. Logistic regression model

logit_q4 <- glm(
  Mortality ~ Age.Group +
    Type.of.Admission +
    Emergency.Department.Indicator +
    APR.Severity.of.Illness.Description +
    APR.Risk.of.Mortality +
    Length.of.Stay,
  data   = train_q4,
  family = binomial
)

summary(logit_q4)


# 6. Predicted probabilities on TEST set 

pred_prob <- predict(logit_q4, newdata = test_q4, type = "response")

# 7. ROC curve and AUC 

roc_q4 <- roc(test_q4$Mortality, pred_prob)
auc_q4 <- auc(roc_q4)

plot(roc_q4,
     col  = "#0066CC",
     lwd  = 2,
     main = paste("ROC Curve for In-Hospital Mortality (AUC =",
                  round(auc_q4, 3), ")"))


# 8. Classify "high-risk" patients at cutoff = 0.10

cutoff <- 0.10
pred_class_num <- ifelse(pred_prob >= cutoff, 1L, 0L)
pred_class_fac <- factor(pred_class_num,
                         levels = c(0, 1),
                         labels = c("Survived", "Died"))

# Confusion matrix on test set
conf_mat_q4 <- confusionMatrix(
  pred_class_fac,
  test_q4$MortalityClass,
  positive = "Died"
)

conf_mat_q4

# Extract key metrics
accuracy    <- conf_mat_q4$overall["Accuracy"]
sensitivity <- conf_mat_q4$byClass["Sensitivity"]
specificity <- conf_mat_q4$byClass["Specificity"]
ppv         <- conf_mat_q4$byClass["Pos Pred Value"]
npv         <- conf_mat_q4$byClass["Neg Pred Value"]

# 9. Visualization: Mortality by APR Risk 

ggplot(train_q4,
       aes(x = APR.Risk.of.Mortality,
           fill = MortalityClass)) +
  geom_bar(position = "fill", color = "black") +
  scale_fill_manual(values = c("Survived" = "#FF9933",
                               "Died"      = "#0066CC")) +
  labs(
    title = "Proportion of In-Hospital Deaths by APR Risk of Mortality",
    x     = "APR Risk of Mortality",
    y     = "Proportion within Risk Group",
    fill  = "Outcome"
  ) +
  theme_bw() +
  theme(
    plot.title   = element_text(hjust = 0.5, face = "bold", size = 14),
    axis.title   = element_text(size = 12),
    axis.text    = element_text(size = 10),
    legend.title = element_text(size = 11),
    legend.text  = element_text(size = 10),
    legend.position = "right"
  )


# 10. Calculate mortality rate for each category of each predictor

mortality_summary <- bind_rows(
  q4_data %>%
    group_by(APR.Risk.of.Mortality) %>%
    summarise(mortality_rate = mean(Mortality)) %>%
    mutate(Factor = "APR Risk of Mortality",
           Category = APR.Risk.of.Mortality),

  q4_data %>%
    group_by(APR.Severity.of.Illness.Description) %>%
    summarise(mortality_rate = mean(Mortality)) %>%
    mutate(Factor = "APR Severity of Illness",
           Category = APR.Severity.of.Illness.Description),

  q4_data %>%
    group_by(Type.of.Admission) %>%
    summarise(mortality_rate = mean(Mortality)) %>%
    mutate(Factor = "Admission Type",
           Category = Type.of.Admission),

  q4_data %>%
    group_by(Age.Group) %>%
    summarise(mortality_rate = mean(Mortality)) %>%
    mutate(Factor = "Age Group",
           Category = Age.Group)
)
mortality_summary$Category <- as.factor(mortality_summary$Category)

# one small chart per factor
ggplot(mortality_summary,
       aes(x = Category, y = mortality_rate * 100, fill = Factor)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ Factor, scales = "free_x") +
  labs(
    title = "Comparison of Mortality Rates Across Predictors",
    x = NULL,
    y = "Mortality Rate (%)"
  ) +
  theme_bw() +
  theme(
    strip.text = element_text(face = "bold", size = 12),
    axis.text.x = element_text(angle = 30, hjust = 1)
  )


# 11. summary

cat("\n--- SMART Question 4 Summary ---\n")
cat("AUC:", round(auc_q4, 3), "\n")
cat("Accuracy:", round(accuracy, 3), "\n")
cat("Sensitivity:", round(sensitivity, 3), "\n")
cat("Specificity:", round(specificity, 3), "\n")
cat("PPV:", round(ppv, 3), "\n")
cat("NPV:", round(npv, 3), "\n")



```


# Results 

## SMART Question1: 
- 


## SMART Question2:
- 


## SMART Question3: 
- 


## SMART Question4: 
- 


# Conclusion & Discussion



# Project Workflow Summary (structured outline)

1. Data acquisition & initial inspection
2. Data cleaning and variable derivation (document every transformation)
3. Exploratory Data Analysis (visual + summary statistics)
4. Hypothesis specification (SMART questions)
5. Statistical testing / modeling (with assumption checks)
6. Interpretation and robustness checks
7. Reporting: visualizations, tables for each SMART question, and final HTML


# Reference

Dataset Link:
https://health.data.ny.gov/Health/Hospital-Inpatient-Discharges-SPARCS-De-Identified/sf4k-39ay/about_data



<!-- End -->
